### 2021-10-16 paper-ancestry-prop.tex
save

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}

%\renewcommand{\arraystretch}{1.8}

\usepackage{xr}
\externaldocument{paper-ancestry-prop-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Using the UK Biobank as a global reference: application to measuring ancestry diversity from GWAS summary statistics}
\author{Florian Priv\'e$^{\text{1}}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$National Centre for Register-based Research, Aarhus University, Aarhus, 8210, Denmark. \\

\noindent Contact:
\begin{itemize}
\item \url{florian.prive.21@gmail.com}
\end{itemize}

\vspace*{5em}

\abstract{
	The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
	Within this dataset, we carefully define 21 distinct ancestry groups from all four corners of the world. 
	Using allele frequencies derived from these global reference groups, we are now able to effectively measure diversity from summary statistics of any genetic dataset.
	Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

[TODO: TALK ABOUT OTHER DATASETS AND WHY CHOOSING THE UKBB]

The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
Here, by carefully using information on self-reported ancestry, country of birth and genetic similarity, we define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations. 
This includes five groups with genetic ancestries from Africa, eight from Europe, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:counts}).
The detailed procedure used to define these groups is presented in the Supplementary Materials.
We provide [TODO FINISH \#VARIANTS].

\begin{table}[h]
	\centering
	\begin{tabular}{|l|c||l|c||l|c|}
		\hline
		Ancestry group & N & Ancestry group & N & Ancestry group & N \\
		\hline
		Japan & 240 (+ 104) & Asia (East) & 504 & Philippines & 304 \\
		Caribbean & 660 & Africa (West) & 655 & Africa (South) & 362 \\
		Africa (East 1) & 193 & Africa (East 2) & 134 & Middle East & 509 \\
		Pakistan & 388 & Sri Lanka & 365 & Bangladesh & 223 (+ 86) \\
		Europe (North West) & 54556 & Finland & 143 (+ 99) & Scandinavia & 342 \\
		Europe (North East) & 259 & Europe (South East) & 344 & Europe (South West) & 605 \\
		Europe (Central) & 559 & Ashkenazi & 1975 & South America & 560 \\
		\hline
	\end{tabular}
	\caption{Number of individuals (N) in each of the 21 ancestry groups we define in the UK Biobank. Numbers in parentheses represent the added individuals from the 1000 Genomes data.}
	\label{tab:counts}
\end{table}

Then, we propose to estimate ancestry proportions from a cohort based on its allele frequencies and the reference set of allele frequencies described previously.
We propose to find the mixture proportions $\alpha_k$ which minimize the following problem:
\(\sum_{j=1}^M \left(w_j  f_j^{(0)} - \sum_{k=1}^K \alpha_k w_j f_j^{(k)} \right)^2 ,\)
where $M$ is the number of variants, $K$ the number of reference populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the cohort of interest.
By further constraining all $w_j$ to be positive and sum to 1 (convex combination) in the optimization, we can interpret these coefficients as ancestry proportions.
This is similar to Summix \cite[]{arriaga2021summix}, although we introduce weights $w_j = 1 / \sqrt{f_j^{(0)} (1 - f_j^{(0)})}$ to account for more variability in more common variants.
This optimization problem is a standard quadratic programming problem that can be solved using R package quadprog \cite[]{turlach2013quadprog}, and that we implement in function [TODO] in our R package bigsnpr \cite[]{prive2017efficient}.

%% Around 320 words already %%

[TODO: CONCLU] Using allele frequencies derived from these groups, we are now able to effectively measure diversity from summary statistics of a genetic dataset.
Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.

\begin{figure}[!htpb]
	%\centerline{\includegraphics[width=0.7\textwidth]{timings.pdf}}
	\caption{Computation times of \texttt{pcadapt} as a function of the number of PCs used. Black points represent timings with pcadapt v4, and the dashed blue line represents the timing with v3, which is independent of the number of PC used. \label{fig:timings}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\section*{Software and code availability}

[TODO]

\section*{Acknowledgements}

F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath) and by a Lundbeck Foundation Fellowship (R335-2019-2339 to Bjarni J. Vilhj\'almsson).
The author thanks GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024.

\section*{Declaration of Interests}

The author declares no competing interests.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
%\section*{Supplementary Materials}
%
%\renewcommand{\thefigure}{S\arabic{figure}}
%\setcounter{figure}{0}
%\renewcommand{\thetable}{S\arabic{table}}
%\setcounter{table}{0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-10-20 paper-ancestry-prop-supp.tex
save

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}

%\renewcommand{\arraystretch}{1.8}

%\usepackage{xr}
%\externaldocument{SCT-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Supplementary Materials}

\date{~ }
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thesection}{S\arabic{section}}
\setcounter{section}{0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Designing a reference of allele frequencies from the UK Biobank}

\subsection{Identifying 26 ancestry groups from the UK Biobank}

In this section, we use the first 16 principal components (PCs, \cite{prive2020efficient}) reported by the UK Biobank (Data-Field 22009, \cite{bycroft2018uk}), the self-reported ancestry (Data-Field 21000), and the country of birth (Data-Field 20115).
For the individuals with a missing country of birth, we assign to them ``United Kingdom'' when they self-identify as ``British'' and ``Ireland'' when they self-identify as ``Irish''.
Hereinafter, we use the Euclidean distance based on the first 16 PCs as we have shown to be an appropriate measure of the genetic distance between populations \cite[]{prive2021high}.
The thresholds we use for this distance are based on carefully inspecting the self-reported ancestry and country of birth of individuals kept or excluded.

Since the country of birth may poorly corresponds to the genetic ancestry that we are interested in, we perform some quality control on this information.
Indeed, there are many individuals born in Africa with an Asian self-reported ancestry; we set their country of birth to missing. 
We also set this information to missing for 1/ individuals born outside of Europe that are too close (log-distance lower than 4) to the UK; 2/ individuals born outside the UK within a log-distance of 2.5 to the UK; 3/ individuals born in Europe not within a log-distance of 5 to the UK.
Finally, we keep only individuals born in South Africa with a log-distance to the UK larger than 6 (individuals of African ancestry). 
After applying these filters, there are still many individuals born outside of the UK, including e.g.\ 3352 individuals from India, 1100 from Germany, 1036 from Nigeria, and 76 countries with at least 50 individuals.

Then, to define a homogeneous group to be used as a reference population, we choose a country, compute the robust PC center of individuals born in this country using the geometric median \cite[]{prive2021high}, and choose to include in this group all individuals within a chosen log-distance to the center. The distance threshold is chosen based on the countries of birth included.
For groups named after a country, we further restrict to individuals born in this country.
We limit the ancestry groups to 2000 individuals in order not to have very uneven sample sizes; we pick at random among the United Kingdom and Ireland groups.
We manually try different country centers and distance thresholds to cover as much of the worldwide populations as possible, without having overlapping groups.
We also include two other groups: one ``Ashkenazi'' group based on a PC center we defined in \cite{prive2021high}, and a ``South America'' group based on individuals with $\text{PC6} > 50$ and $\text{PC8} < -30$. 
Finally, we investigate individuals that are far away from any of the groups defined previously, and add a last group based on a center from India.

[TODO: OUTPUT COUNTRIES FROM EACH GROUP]

\subsection{Computing allele frequencies}

We download the 1000 Genomes (1KG) Project data \cite[]{10002015global}. 
Using PLINK \cite[]{chang2015second}, we remove 9 outlier individuals identified in \cite{martin2017human}, filter for founders, non-sex chromosomes, non-multiallelic variants with minor allele counts of at least 10.
From the UK Biobank, we select imputed variants with a minor allele frequency larger than 0.001 and INFO score larger than 0.3, and match these variants with the ones from the filtered 1KG data described previously.
We identify 12,373,666 genetic variants in common, for which we compute allele frequencies for each of the 26 1KG populations. 
We also compute allele frequencies from the BGEN imputed data for each of the 26 UK Biobank groups we identified previously.

\subsection{Designing the final sets of allele frequencies}

We select variants with an INFO score of at least 0.5 in all 26 UK Biobank ancestry groups. [TODO: NUMBER]
Based on the remaining variants, we compute the overall fixation indices $F_{ST}$ between the UK Biobank groups and the 1KG populations.
For each of the 1KG populations, we identify the closest ancestry group among those we defined, compute the $F_{ST}$ for all variants and scale them by dividing by the overall $F_{ST}$ with this group.
For each variant, we then sum these scaled $F_{ST}$ (26 different ones corresponding to the UK Biobank ancestry groups), and identify and remove 18,344 variants with a large score.

Finally, for the Finland, Bangladesh and Japan ancestry groups, we compute a weighted average of allele frequencies with the ones from the 1KG data (FIN, BEB and JPT) to increase the small sample size of these groups.

[TODO: REPORT FREQUENCIES? FST?]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!htpb]
	%\centerline{\includegraphics[width=0.75\textwidth]{compare-with-old.png}}
	\caption{For one dataset with a large number of missing values, chi-square statistics from pcadapt v4.1.0 (y-axis) versus statistics from v3.0.4 (x-axis). Correlation between these statistics is of 99.76\%. \label{fig:corstats}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-10-20 paper-ancestry-prop.tex
save

@@
 	The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
-	Within this dataset, we carefully define 21 distinct ancestry groups from all four corners of the world. 
+	Within this dataset, we carefully define 26 distinct ancestry groups from all four corners of the world. 
 	Using allele frequencies derived from these global reference groups, we are now able to effectively measure diversity from summary statistics of any genetic dataset.
@@
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
-Here, by carefully using information on self-reported ancestry, country of birth and genetic similarity, we define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-This includes five groups with genetic ancestries from Africa, eight from Europe, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:counts}).
+Here, by carefully using information on self-reported ancestry, country of birth and genetic similarity, we define 26 distinct ancestry groups from the UK Biobank to be used as global reference populations.
+This includes ten groups with genetic ancestries from Europe, seven from Africa, four from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:counts}).
 The detailed procedure used to define these groups is presented in the Supplementary Materials.
@@
 where $M$ is the number of variants, $K$ the number of reference populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the cohort of interest.
-By further constraining all $w_j$ to be positive and sum to 1 (convex combination) in the optimization, we can interpret these coefficients as ancestry proportions.
-This is similar to Summix \cite[]{arriaga2021summix}, although we introduce weights $w_j = 1 / \sqrt{f_j^{(0)} (1 - f_j^{(0)})}$ to account for more variability in more common variants.
-This optimization problem is a standard quadratic programming problem that can be solved using R package quadprog \cite[]{turlach2013quadprog}, and that we implement in function [TODO] in our R package bigsnpr \cite[]{prive2017efficient}.
+By further constraining all $\alpha_k$ to be positive and sum to 1 (convex combination) in the optimization, we can interpret these coefficients as ancestry proportions.
+This is similar to Summix \cite[]{arriaga2021summix}, with the difference that we introduce weights $w_j = 1 / \sqrt{f_j^{(0)} (1 - f_j^{(0)})}$ to account for more variability in more common variants.
+This optimization problem is a standard quadratic programming problem that can be solved using R package quadprog \cite[]{turlach2013quadprog}, and that we implement in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}.
 
@@
 
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{lrccccccccccccc}
+		\hline
+		Population & N & BBJ & body-fat & BrCa & CAD & covid & eczema & epilepsy & FinnGen & GERA & kidney & PAGE & PrCa & UKBB \\
+		\hline
+		Africa (East 1) & 193 & & 0.1 &&0.5 & 0.4 & 0.1 & & & & 1 & 0.8 &&0.1 \\
+		Africa (East 2) & 134 & &0.3& & & 0.1 & & & & 0.7 & 0.4 & & \\
+		Africa (South) & 362 & & 0.9 & & 0.5 & 0.7 & 0.1 & 0.1 & & 0.3 & 1.2 & 4.8 & 0.1 & 0.2 \\
+		Africa (West) & 655 & & & 0.2 & 1.2 & 2.1 & 0.1 & 0.4 & & 1 & & 14.1 & & 0.7\\
+		\hline
+		Ashkenazi & 1975 & & 1.3 & 0.6 & 1.4 & 0.5 & 1.1 & 2.6 & & 5.2 & 0.9 & 1 & 2.5 & 1 \\
+		Asia (East) & 504 & 4.3 & 2.6 & 0.8 & 3.2 & 0.4 & 0.2 & 2.9 & & 3.7 & 2.6 & 2.8 & 0.5 & 0.4 \\
+		Bangladesh & 223 & & 2.6 & 1.1 & 2.7 & 0.9 & & & & 0.2 & 0.8 & 1.3 & 0.8 & 0.5 \\
+		Caribbean & 660 & & 1.4 & 0.1 & 1.5 & 2.3 & 0.3 & 0.4 & & 1.3 & 0.5 & 14.5 & 0.2 & 0.9 \\
+		Europe (Central) & 559 & & 8 & 5.9 & 5.7 & 5.5 & 5.8 & 5.1 & 1.7 & 4.2 & 2.5 & 0.5 & 5.6 & \\
+		\hline
+		Europe (North East) & 259 & & 1.9 & 1.3 & 3.1 & 5.1 & 2.3 & 2 & 7 & 1.4 & 1 & 0.8 & 2.4 & 0.6 \\
+		Europe (North West) & 54556 & & 51.2 & 62.6 & 49.1 & 56.1 & 67.4 & 69.8 & & 58.1 & 66.6 & 6 & 58.8 & 92.8 \\
+		Europe (South East) & 344 & & 4.4 & 3 & 3.5 & 1.7 & 3.3 & 4.2 & & 3.7 & 1.4 & 0.6 & 3.2 & \\
+		\hline
+		Europe (South West) & 605 & & 2.2 & 3.4 & 1.7 && 2.5 & 4.2 & & 5.9 & 1 & 9.3 & 4.1 & 0.5 \\
+		Finland & 143 & & 7.5 & 8.3 & 5.8 & 12.7 & 10.2 & 4.7 & 83.2 & 1.1 & 4.1 & & 11.6 & \\
+		Japan & 240 & 95.7 & 2.5 & 0.2 & 1.6 & 0.5 & 0.6 &&0.4 & 1.6 & 11.9 & 8.4 & 0.2 & \\
+		Middle East & 509 & & 1.3 & 0.9 & 4 & 0.5 & 0.6 & 2.1 & & 2.3 & 0.3 & & 1.1 & 0.4 \\
+		Pakistan & 388 & & 2.7 & 0.4 & 6.2 & 1 & & & & & 0.1 & & & 1.2 \\
+		\hline
+		Philippines & 304 & & & & & 0.1 & & 0.4 & & 1.8 & 0.6 & 4 & & 0.1 \\
+		Scandinavia & 342 & & 6.1 & 9 & 4.4 & 6.6 & 4.6 & 0.8 & 7.6 & 2.7 & 1.7 & & 7.5 & \\
+		South America & 560 & & 1.1 & 1.1 & 1.7 & 2.3 & 0.6 & 0.3 & 0.1 & 5.5 & 0.5 & 30.1 & 0.8 & 0.2 \\
+		Sri Lanka & 365 & & 1.9 & 1 & 2.4 & 0.4 & & & & 0.1 & 0.6 & 0.7 & 0.5 & 0.4 \\
+		\end{tabular}
+\end{table}
+
 [TODO: CONCLU] Using allele frequencies derived from these groups, we are now able to effectively measure diversity from summary statistics of a genetic dataset.
 Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
-
-\begin{figure}[!htpb]
-	%\centerline{\includegraphics[width=0.7\textwidth]{timings.pdf}}
-	\caption{Computation times of \texttt{pcadapt} as a function of the number of PCs used.
-Black points represent timings with pcadapt v4, and the dashed blue line represents the timing with v3, which is independent of the number of PC used.
-\label{fig:timings}}
-\end{figure}
 

### 2021-10-20 paper-ancestry-prop.tex
save

@@
 
-\vspace*{5em}
+\vspace*{3em}
 
@@
 
+\vspace*{5em}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\clearpage
+%\clearpage
 
-[TODO: TALK ABOUT OTHER DATASETS AND WHY CHOOSING THE UKBB]
-
+% 185 words
+Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project, the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
+However, these datasets are not very large and do not provide many individuals per population.
 The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
-Here, by carefully using information on self-reported ancestry, country of birth and genetic similarity, we define 26 distinct ancestry groups from the UK Biobank to be used as global reference populations.
+Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 26 distinct ancestry groups from the UK Biobank to be used as global reference populations.
 This includes ten groups with genetic ancestries from Europe, seven from Africa, four from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:counts}).
@@
 
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|l|c||l|c||l|c|}
-		\hline
-		Ancestry group & N & Ancestry group & N & Ancestry group & N \\
-		\hline
-		Japan & 240 (+ 104) & Asia (East) & 504 & Philippines & 304 \\
-		Caribbean & 660 & Africa (West) & 655 & Africa (South) & 362 \\
-		Africa (East 1) & 193 & Africa (East 2) & 134 & Middle East & 509 \\
-		Pakistan & 388 & Sri Lanka & 365 & Bangladesh & 223 (+ 86) \\
-		Europe (North West) & 54556 & Finland & 143 (+ 99) & Scandinavia & 342 \\
-		Europe (North East) & 259 & Europe (South East) & 344 & Europe (South West) & 605 \\
-		Europe (Central) & 559 & Ashkenazi & 1975 & South America & 560 \\
-		\hline
-	\end{tabular}
-	\caption{Number of individuals (N) in each of the 21 ancestry groups we define in the UK Biobank. Numbers in parentheses represent the added individuals from the 1000 Genomes data.}
-	\label{tab:counts}
-\end{table}
-
-Then, we propose to estimate ancestry proportions from a cohort based on its allele frequencies and the reference set of allele frequencies described previously.
+% 185 words
+Then, as a use-case for this new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only.
 We propose to find the mixture proportions $\alpha_k$ which minimize the following problem:
@@
 
-[TODO: CONCLU] Using allele frequencies derived from these groups, we are now able to effectively measure diversity from summary statistics of a genetic dataset.
+Citations for the allele frequencies used: the UK Biobank (UKBB, \cite{bycroft2018uk}), the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, kidney function \cite[]{yu2021polygenic}, COVID-19 \cite[]{ganna2021mapping}, epilepsy \cite[]{consortium2018genome}, body fat percentage \cite[]{lu2016new}, eczema \cite[]{paternoster2015multi}, prostate cancer (PrCa, \cite{schumacher2018association}), breast cancer (BrCa, \cite{michailidou2017association}), and coronary artery disease (CAD, \cite{nikpay2015comprehensive}).
+Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
+
+% 83 words
+Here we provide an unprecedentedly large and diverse reference set of allele frequencies using the UK Biobank.
+We then show how to effectively measure diversity from GWAS summary statistics using this new set of reference allele frequencies.
 Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
+This is one of the many applications of the new set we provide.
 
@@
 
-\clearpage
+%\clearpage
 

### 2021-10-22 code/1-prepare-ukbb-groups.R
export code and results

@@ -1,0 +1,265 @@
+library(dplyr)
+library(bigreadr)
+
+#### Get PC / ancestry info from UKBB ####
+
+# file.symlink("~/NCRR-PRS/faststorage/UKBB/", ".")
+
+df_UKBB <- runonce::save_run({
+
+  # Self-reported ancestry
+  code_ancestry <- fread2("UKBB/coding1001.tsv")
+  # Country of birth
+  code_country   <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+  code_continent <- filter(fread2("UKBB/coding89.tsv"), selectable == "N")
+
+  eid_imputed <- fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+
+  fread2(
+    "UKBB/ukb41181.csv",
+    select = c("eid", "21000-0.0", "20115-0.0", paste0("22009-0.", 1:16)),
+    col.names = c("eid", "ancestry", "country", paste0("PC", 1:16))
+  ) %>%
+    filter(eid %in% eid_imputed) %>%
+    mutate(
+      ancestry = factor(ancestry, levels = code_ancestry$coding,
+                        labels = code_ancestry$meaning) %>%
+        forcats::fct_recode("Other White" = "Any other white background",
+                            "Other Asian" = "Any other Asian background",
+                            "Other Asian" = "Asian or Asian British",
+                            "Other Black" = "Any other Black background",
+                            "Other Black" = "Black or Black British") %>%
+        forcats::fct_other(drop = c("Prefer not to answer", "Do not know", "Mixed",
+                                    "Other ethnic group", "Any other mixed background",
+                                    "White and Asian", "White and Black Caribbean",
+                                    "White and Black African"),
+                           other_level = NA) %>%
+        droplevels(exclude = NA),
+
+      continent = factor(country, levels = code_country$coding,
[…]

### 2021-10-22 code/2-prepare-1000G-freq.R

@@ -1,0 +1,74 @@
+# https://www.cog-genomics.org/plink/2.0/resources#1kg_phase3
+pgen_zst <- runonce::download_file(
+  "https://www.dropbox.com/s/afvvf1e15gqzsqo/all_phase3.pgen.zst?dl=1",
+  dir = "tmp-data")
+pvar_zst <- runonce::download_file(
+  "https://www.dropbox.com/s/op9osq6luy3pjg8/all_phase3.pvar.zst?dl=1",
+  dir = "tmp-data")
+psam <- runonce::download_file(
+  "https://www.dropbox.com/s/yozrzsdrwqej63q/phase3_corrected.psam?dl=1",
+  dir = "tmp-data", fname = "all_phase3.psam")
+
+library(bigsnpr)
+plink2 <- download_plink2("tmp-data")
+
+sysglue <- function(...) system(glue::glue(...))
+pgen <- sub('\\.zst', '', pgen_zst)
+if (!file.exists(pgen)) sysglue("{plink2} --zst-decompress {pgen_zst} > {pgen}")
+pvar <- sub('\\.zst', '', pvar_zst)
+if (!file.exists(pvar)) sysglue("{plink2} --zst-decompress {pvar_zst} > {pvar}")
+
+# https://dx.doi.org/10.1016%2Fj.ajhg.2017.03.004
+outliers <- c("NA20314", "HG00731", "HG00732", "HG01880", "HG01882",
+              "HG01944", "HG02497", "NA20320", "NA20321")
+bigsnpr:::write.table2(cbind.data.frame(0, outliers), (tmp <- tempfile()))
+
+sysglue("{plink2} --pfile tmp-data/all_phase3",
+        " --remove {tmp} --keep-founders",
+        " --max-alleles 2 --mac 10 --autosome",
+        " --make-bed --out tmp-data/all_phase3",
+        " --threads {nb_cores()[[1]]} --memory 50e3")
+
+obj.bed <- bed("tmp-data/all_phase3.bed")
+
+library(doParallel)
+cl <- makeCluster(22)
+snp_info <- do.call("rbind", parLapply(cl, 1:22, function(chr) {
+  library(dplyr)
+  paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt") %>%
+    bigreadr::fread2(showProgress = FALSE) %>%
+    filter(V6 > 0.001, V8 > 0.3) %>%
[…]

### 2021-10-22 code/3-prepare-ukbb-freq.R

@@ -1,0 +1,64 @@
+data_1kg <- readRDS("tmp-data/info_af_1kg.rds")
+## Get all SNP IDs and split them in blocks
+all_snp_id <- with(data_1kg, split(paste(chr, pos, a0, a1, sep = "_"), as.factor(1:22)[chr]))
+lengths(all_snp_id)
+#      1       2       3       4       5       6       7       8       9      10      11
+# 981656 1036649  878387  899012  782944  819453  724662  674884  530864  627816  610113
+#     12      13      14      15      16      17      18      19      20      21      22
+# 598814  448686  404642  358726  379725  346916  352132  293684  275652  172731  175518
+
+library(dplyr)
+grid <- purrr::map_dfr(1:22, function(chr) {
+  bgenfile <- paste0("UKBB/bgen/ukb_imp_chr", chr, "_v3.bgen")
+  all_splits <- bigparallelr::split_vec(all_snp_id[[chr]], block_len = 50e3)
+  tibble::tibble(bgenfile, snp_id = all_splits)
+})
+nrow(grid) # 257 parts
+
+all_eid <- readRDS("tmp-data/info-UKBB.rds")$eid
+eid_sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+all_ind <- lapply(readRDS("data/list_ind_pop.rds"), function(ind) {
+  match(all_eid[ind], eid_sample)
+})
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(workers = print(nrow(grid)), resources = list(
+  t = "12:00:00", c = NCORES, mem = "120g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("tmp-data/bgen")
+bigassertr::assert_dir("all_freq_ukbb")
+bigassertr::assert_dir("log")
+
+furrr::future_pwalk(mutate(grid, id = row_number()), function(bgenfile, snp_id, id) {
+
+  for (ipop in seq_along(all_ind)) {
+
+    res_file <- paste0("all_freq_ukbb/pop", ipop, "_part", id, ".rds")
+
+    runonce::save_run({
+
+      tmp <- tempfile(tmpdir = "tmp-data/bgen")
+      on.exit(file.remove(paste0(tmp, c(".bk", ".rds"))), add = TRUE)
+
+      library(bigsnpr)
[…]

### 2021-10-22 code/4-filter-merge-freq.R

@@ -1,0 +1,131 @@
+data_1kg <- readRDS("tmp-data/info_af_1kg.rds")
+data_ukbb <- readRDS("tmp-data/info_af_ukbb.rds")
+
+INFO <- do.call(cbind, lapply(data_ukbb, function(.) .$info))
+round(sort(colMeans(INFO, na.rm = TRUE)), 3)
+# Japan         Asia (East)         Philippines          Bangladesh           Sri Lanka
+# 0.628               0.659               0.680               0.748               0.752
+# Pakistan    Africa (Central)      Africa (South)       Africa (West)     Africa (East 1)
+#    0.768               0.785               0.792               0.799               0.811
+# Africa (East 2) Europe (South East) Europe (North East)               India               Italy
+#           0.813               0.815               0.820               0.820               0.821
+# Europe (Central)             Ireland         Scandinavia         Middle East           Ashkenazi
+#            0.822               0.825               0.835               0.835               0.846
+# United Kingdom           Caribbean      Africa (North) Europe (South West)             Finland
+#          0.856               0.869               0.874               0.876               0.878
+# South America
+#         0.909
+
+mean_info <- rowMeans(INFO)
+min_info <- matrixStats::rowMins(INFO)
+hist(mean_info)
+hist(min_info)
+ind_keep <- which(min_info > 0.4)
+length(ind_keep) # 5,840,630
+
+X_ukbb <- do.call(cbind, lapply(data_ukbb, function(.) .$freq[ind_keep]))
+summary(rowMeans(X_ukbb))
+#     Min.  1st Qu.   Median     Mean  3rd Qu.     Max.
+# 0.004086 0.136399 0.274173 0.335182 0.497979 0.992190
+X_1kg <- as.matrix(data_1kg[ind_keep, -(1:5)])
+
+pop_1kg_size  <- readRDS("tmp-data/pop_1kg_size.rds")
[…]

### 2021-10-22 code/5-assess-ancestry-in-sumstats.R

@@ -1,0 +1,316 @@
+library(bigreadr)
+library(bigsnpr)
+library(dplyr)
+bigassertr::assert_dir("res_ancestry")
+
+all_freq <- readRDS("data/all_freq.rds")
+# bigreadr::fwrite2(mutate_at(all_freq, -(1:5), ~ round(., 5)), "ref_freqs.csv.gz")
+
+snp_match_ancestry <- function(sumstats, join_by_pos = TRUE) {
+
+  matched <- bigsnpr::snp_match(
+    transform(sumstats, chr = as.integer(chr), beta = 1),
+    all_freq[1:5],
+    join_by_pos = join_by_pos,
+    match.min.prop = 0.1
+  ) %>%
+    mutate(freq = ifelse(beta > 0, freq, 1 - freq))
+
+  X <- as.matrix(all_freq[matched$`_NUM_ID_`, -(1:5)])
+  y <- matched$freq
+  snp_ancestry_summary(y, X)
+}
+
+
+## UKBB
+
+ukbb <- bed("../paper-ancestry-matching/data/ukbb.bed")
+af_UKBB <- bed_MAF(ukbb, ncores = nb_cores())$af
+sumstats <- transmute(ukbb$map, chr = chromosome, pos = physical.pos,
+                      a0 = allele2, a1 = allele1, freq = af_UKBB)
+str(res <- snp_match_ancestry(sumstats))
+saveRDS(res, "res_ancestry/UKBB.rds")
+
+
+## Caribbean defined previously
+
+sumstats <- cbind.data.frame(
+  readRDS("tmp-data/info_af_1kg.rds")[1:5],
+  readRDS("tmp-data/info_af_ukbb.rds")$Caribbean
+) %>%
+  filter(info > 0.4)
+
+str(res <- snp_match_ancestry(sumstats))
+saveRDS(res, "res_ancestry/caribbean.rds")
+
+
+
+## AMR in 1KG
+
+obj.bed <- bed(download_1000G("tmp-data"))
+fam2 <- bigreadr::fread2(paste0(obj.bed$prefix, ".fam2"))
+ind_row <- which(fam2$`Super Population` == "AMR")
+
+af <- bed_MAF(obj.bed, ind_row, ncores = nb_cores())$af
+sumstats <- dplyr::transmute(obj.bed$map, chr = chromosome, pos = physical.pos,
+                             a1 = allele1, a0 = allele2, freq = af)
+
[…]

### 2021-10-24 paper-ancestry-prop-supp.tex
save

@@
 We manually try different country centers and distance thresholds to cover as much of the worldwide populations as possible, without having overlapping groups.
-We also include two other groups: one ``Ashkenazi'' group based on a PC center we defined in \cite{prive2021high}, and a ``South America'' group based on individuals with $\text{PC6} > 50$ and $\text{PC8} < -30$.
+We also include two other groups: one ``Ashkenazi'' group based on a PC center we defined in \cite{prive2021high}, and a ``South America'' group based on individuals with $\text{PC6} > 50$ and $\text{PC8} < -30$ (clearly separated from the rest of individuals from the UKBB).
 Finally, we investigate individuals that are far away from any of the groups defined previously, and add a last group based on a center from India.
 
-[TODO: OUTPUT COUNTRIES FROM EACH GROUP]
+For each of the 26 ancestry groups we define, we list the individuals' countries of birth (with at least 5 individuals, where ``NA'' means unknown):
+\begin{itemize}
+\item \textbf{Japan:~~} Japan: 240
+\item \textbf{Asia (East):~~} Hong Kong: 390 --- Malaysia: 179 --- NA: 122 --- China: 114 --- Singapore: 58 --- Vietnam: 33 --- Indonesia: 14 --- Taiwan: 10 --- Thailand: 8 --- Brunei: 6 --- Macau (Macao): 6
+\item \textbf{Philippines:~~} Philippines: 295
+\item \textbf{Africa (South):~~} Zimbabwe: 251 --- Uganda: 57 --- South Africa: 47 --- Zambia: 41 --- Malawi: 10 --- NA: 10 --- Tanzania: 6 --- Congo: 5
+\item \textbf{Africa (North):~~} Algeria: 69 --- Morocco: 62 --- Egypt: 54 --- Libya: 36 --- NA: 18 --- Tunisia: 13
+\item \textbf{Africa (East 1):~~} Somalia: 81 --- Ethiopia: 58 --- Sudan: 54 --- Eritrea: 46 --- NA: 27
+\item \textbf{Africa (East 2):~~} Kenya: 104 --- Uganda: 60 --- Burundi: 18 --- NA: 16 --- Sudan: 14 --- Tanzania: 14 --- Rwanda: 13 --- Nigeria: 8 --- Somalia: 7 --- Angola: 5 --- Zimbabwe: 5
+\item \textbf{Caribbean:~~} Caribbean: 327 --- NA: 310 --- Barbados: 73 --- The Guianas: 23 --- Antigua and Barbuda: 7
+\item \textbf{Africa (West):~~} Nigeria: 217 --- Ghana: 175 --- NA: 122 --- Sierra Leone: 98 --- Caribbean: 80 --- Ivory Coast: 7 --- Liberia: 7 --- The Guianas: 7 --- Togo: 6
+\item \textbf{Africa (Central):~~} Congo: 117 --- Cameroon: 42 --- Nigeria: 33 --- NA: 33 --- Angola: 21 --- Caribbean: 8 --- The Guianas: 7
+\item \textbf{Middle East:~~} Iraq: 240 --- Iran: 172 --- Turkey: 55 --- NA: 31 --- Syria: 11
+\item \textbf{United Kingdom:~~} United Kingdom: 2000
+\item \textbf{Ireland:~~} Ireland: 2000
+\item \textbf{Finland:~~} Finland: 143
+\item \textbf{Scandinavia:~~} Denmark: 142 --- Norway: 82 --- Sweden: 75 --- NA: 68 --- Germany: 25 --- Netherlands: 17 --- Iceland: 5
+\item \textbf{Europe (South West):~~} Spain: 279 --- Portugal: 198 --- NA: 91 --- France: 25 --- Gibraltar: 7
+\item \textbf{Italy:~~} Italy: 345
+\item \textbf{Europe (South East):~~} NA: 115 --- Romania: 47 --- Serbia/Montenegro: 41 --- Bosnia and Herzegovina: 34 --- Bulgaria: 33 --- Croatia: 28 --- Hungary: 17 --- Italy: 11 --- Macedonia: 8
+\item \textbf{Europe (Central):~~} NA: 196 --- Germany: 111 --- Czech Republic: 79 --- Poland: 50 --- Austria: 33 --- Hungary: 27 --- Slovakia: 25 --- Croatia: 10 --- Slovenia: 8 --- France: 7
+\item \textbf{Europe (North East):~~} Russia: 88 --- NA: 87 --- Poland: 72 --- Ukraine: 34 --- Latvia: 10 --- Germany: 7 --- Kazakhstan: 6 --- Lithuania: 5
+\item \textbf{Pakistan:~~} Pakistan: 400
+\item \textbf{Sri Lanka:~~} Sri Lanka: 372
+\item \textbf{Bangladesh:~~} Bangladesh: 223
+\item \textbf{South America:~~} Colombia: 173 --- Chile: 57 --- Mexico: 51 --- Peru: 50 --- Ecuador: 33 --- Bolivia: 21 --- NA: 20 --- Venezuela: 15 --- Brazil: 10 --- United Kingdom: 8 --- USA: 8 --- Argentina: 5
+\item \textbf{Ashkenazi:~~} United Kingdom: 1182 --- NA: 572 --- USA: 115 --- Israel: 27 --- Hungary: 12 --- Ireland: 8 --- France: 7 --- Canada: 6 --- Russia: 5
+\item \textbf{India:~~} NA: 258 --- India: 108 --- Sri Lanka: 16 --- Pakistan: 11
+\end{itemize}
 
@@
 
-We select variants with an INFO score of at least 0.5 in all 26 UK Biobank ancestry groups. [TODO: NUMBER]
-Based on the remaining variants, we compute the overall fixation indices $F_{ST}$ between the UK Biobank groups and the 1KG populations.
+We select variants with an INFO score of at least 0.4 in all 26 UK Biobank ancestry groups.
+Based on the remaining 5,840,630 variants, we compute the overall fixation indices $F_{ST}$ between the UK Biobank groups and the 1KG populations (See the description of the 1KG populations at \url{https://www.coriell.org/1/NHGRI/Collections/1000-Genomes-Collections/1000-Genomes-Project}).
+This enable us to check that the group we defined appropriate capture the ancestries they should.
+We report all $F_{ST}$ in a CSV file, and show only the lowest for each of the 26 ancestry groups we define here:
+\begin{itemize}
+\item \textbf{Japan:~~} JPT: 0.00033 --- CHB: 0.007 --- CHS: 0.0089 --- Asia (East): 0.01 --- KHV: 0.014 --- CDX: 0.017 --- Philippines: 0.021
+\item \textbf{Asia (East):~~} CHS: 0.0009 --- KHV: 0.0021 --- CHB: 0.0029 --- CDX: 0.0034 --- Philippines: 0.01 --- Japan: 0.01 --- JPT: 0.011
+\item \textbf{Philippines:~~} Asia (East): 0.01 --- KHV: 0.01 --- CDX: 0.011 --- CHS: 0.012 --- CHB: 0.015 --- Japan: 0.021 --- JPT: 0.021
+\item \textbf{Africa (South):~~} Africa (Central): 0.0017 --- LWK: 0.0035 --- Caribbean: 0.0043 --- Africa (West): 0.0044 --- Africa (East 2): 0.005 --- YRI: 0.005 --- ESN: 0.0055
+\item \textbf{Africa (North):~~} PUR: 0.01 --- Middle East: 0.01 --- Italy: 0.01 --- Europe (South West): 0.011 --- TSI: 0.012 --- IBS: 0.012 --- Ashkenazi: 0.012
+\item \textbf{Africa (East 1):~~} Africa (North): 0.018 --- ASW: 0.027 --- Africa (East 2): 0.028 --- PUR: 0.029 --- Middle East: 0.036 --- ACB: 0.037 --- CLM: 0.039
+\item \textbf{Africa (East 2):~~} LWK: 0.0033 --- Caribbean: 0.0045 --- ASW: 0.0048 --- ACB: 0.0048 --- Africa (South): 0.005 --- Africa (Central): 0.0056 --- Africa (West): 0.0074
+\item \textbf{Caribbean:~~} ACB: 0.00048 --- Africa (West): 0.0011 --- YRI: 0.0022 --- Africa (Central): 0.0023 --- ESN: 0.0028 --- ASW: 0.003 --- MSL: 0.0042
+\item \textbf{Africa (West):~~} YRI: 0.00061 --- Caribbean: 0.0011 --- ESN: 0.0015 --- Africa (Central): 0.0019 --- ACB: 0.0023 --- MSL: 0.0028 --- Africa (South): 0.0044
+\item \textbf{Africa (Central):~~} Africa (South): 0.0017 --- Africa (West): 0.0019 --- Caribbean: 0.0023 --- YRI: 0.0024 --- ESN: 0.0028 --- ACB: 0.0036 --- LWK: 0.0045
+\item \textbf{Middle East:~~} Italy: 0.0043 --- TSI: 0.0059 --- Europe (South East): 0.007 --- Ashkenazi: 0.007 --- Europe (South West): 0.008 --- India: 0.0088 --- IBS: 0.0089
+\item \textbf{United Kingdom:~~} Scandinavia: 0.00043 --- Ireland: 0.00069 --- CEU: 0.001 --- GBR: 0.0011 --- Europe (Central): 0.0012 --- Europe (South East): 0.0021 --- Europe (South West): 0.0023
+\item \textbf{Ireland:~~} United Kingdom: 0.00069 --- Scandinavia: 0.0015 --- GBR: 0.0017 --- CEU: 0.0018 --- Europe (Central): 0.0023 --- Europe (South West): 0.0033 --- Europe (South East): 0.0033
+\item \textbf{Finland:~~} FIN: 0.00041 --- Europe (North East): 0.0052 --- Scandinavia: 0.0054 --- Europe (Central): 0.0056 --- United Kingdom: 0.0066 --- CEU: 0.0067 --- GBR: 0.0072
+\item \textbf{Scandinavia:~~} United Kingdom: 0.00043 --- Europe (Central): 0.0012 --- CEU: 0.0013 --- Ireland: 0.0015 --- GBR: 0.0015 --- Europe (South East): 0.0026 --- Europe (North East): 0.0027
+\item \textbf{Europe (South West):~~} IBS: 0.00079 --- Italy: 0.0014 --- Europe (South East): 0.002 --- TSI: 0.0021 --- United Kingdom: 0.0023 --- Europe (Central): 0.003 --- CEU: 0.003
+\item \textbf{Italy:~~} TSI: 0.0011 --- Europe (South West): 0.0014 --- Europe (South East): 0.0017 --- IBS: 0.0023 --- Europe (Central): 0.004 --- United Kingdom: 0.0041 --- Middle East: 0.0043
+\item \textbf{Europe (South East):~~} Europe (Central): 0.00075 --- Italy: 0.0017 --- Europe (North East): 0.0017 --- Europe (South West): 0.002 --- United Kingdom: 0.0021 --- Scandinavia: 0.0026 --- TSI: 0.0027
+\item \textbf{Europe (Central):~~} Europe (North East): 0.00053 --- Europe (South East): 0.00075 --- Scandinavia: 0.0012 --- United Kingdom: 0.0012 --- CEU: 0.0019 --- GBR: 0.0023 --- Ireland: 0.0023
+\item \textbf{Europe (North East):~~} Europe (Central): 0.00053 --- Europe (South East): 0.0017 --- Scandinavia: 0.0027 --- United Kingdom: 0.003 --- CEU: 0.0036 --- Ireland: 0.0039 --- GBR: 0.0039
+\item \textbf{Pakistan:~~} PJL: 0.0026 --- India: 0.0041 --- GIH: 0.0053 --- Bangladesh: 0.0059 --- Sri Lanka: 0.0065 --- ITU: 0.0068 --- STU: 0.0074
+\item \textbf{Sri Lanka:~~} STU: 0.00034 --- ITU: 0.0015 --- Bangladesh: 0.0023 --- BEB: 0.0025 --- PJL: 0.0037 --- GIH: 0.0043 --- Pakistan: 0.0065
+\item \textbf{Bangladesh:~~} BEB: 0.00068 --- Sri Lanka: 0.0023 --- STU: 0.0028 --- ITU: 0.003 --- PJL: 0.0037 --- GIH: 0.0044 --- Pakistan: 0.0059
+\item \textbf{South America:~~} MXL: 0.0023 --- CLM: 0.0053 --- PUR: 0.014 --- PEL: 0.019 --- India: 0.022 --- Pakistan: 0.027 --- Europe (South East): 0.029
+\item \textbf{Ashkenazi:~~} Italy: 0.0051 --- TSI: 0.0064 --- Europe (South West): 0.0069 --- Middle East: 0.007 --- Europe (South East): 0.0072 --- IBS: 0.008 --- Europe (Central): 0.0097
+\item \textbf{India:~~} Pakistan: 0.0041 --- PJL: 0.007 --- Middle East: 0.0088 --- Europe (South East): 0.0091 --- Bangladesh: 0.0095 --- Europe (Central): 0.0096 --- United Kingdom: 0.0097
+\end{itemize}
+
 For each of the 1KG populations, we identify the closest ancestry group among those we defined, compute the $F_{ST}$ for all variants and scale them by dividing by the overall $F_{ST}$ with this group.
 For each variant, we then sum these scaled $F_{ST}$ (26 different ones corresponding to the UK Biobank ancestry groups), and identify and remove 18,344 variants with a large score.
+This enable us to identify variants with possible errors in the UK Biobank.
 
 Finally, for the Finland, Bangladesh and Japan ancestry groups, we compute a weighted average of allele frequencies with the ones from the 1KG data (FIN, BEB and JPT) to increase the small sample size of these groups.
+Since some ancestry groups are genetically very close to each other (very low $F_{ST}$), we merge them by averaging their allele frequencies.
+Thus we merge the ``Irish'', ``United Kingdom'' and ``Scandinavia'' groups  to form a new Northwestern European group.
+We also merge the groups from Central, North-East and South-East Europe to form a new Eastern European group.
+We remove the ``India'' group since the $F_{ST}$ with other populations seem to place this group north of India, instead of in the middle of India, as we sought.
+We also remove the ``Caribbean'' group since it is an admixed population (estimated by our method to be of 92.2\% African, 6.2\% Northwestern European and 1\% Southwestern European ancestries, see Table 1 of main text).
 
-[TODO: REPORT FREQUENCIES? FST?]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[!htpb]
-	%\centerline{\includegraphics[width=0.75\textwidth]{compare-with-old.png}}
-	\caption{For one dataset with a large number of missing values, chi-square statistics from pcadapt v4.1.0 (y-axis) versus statistics from v3.0.4 (x-axis).
-Correlation between these statistics is of 99.76\%.
-\label{fig:corstats}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+This results in the set of 5,816,590 variants for 20 populations, which we provide for download, and that we use to estimate ancestry proportions in GWAS summary statistics.
 

### 2021-10-24 paper-ancestry-prop.tex
save

--- 2021-10-16 paper-ancestry-prop.tex
@@
 
+\usepackage{rotating}
+
 \usepackage{dsfont}
@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 	The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
-	Within this dataset, we carefully define 21 distinct ancestry groups from all four corners of the world. 
+	Within this dataset, we carefully define 20 distinct ancestry groups from all four corners of the world. 
 	Using allele frequencies derived from these global reference groups, we are now able to effectively measure diversity from summary statistics of any genetic dataset.
@@
 }
+
+\vspace*{5em}
 
@@
 
-[TODO: TALK ABOUT OTHER DATASETS AND WHY CHOOSING THE UKBB]
-
+Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project, the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
+However, these datasets do not contain many individuals per population and therefore are large enough to provide accurate allele frequency estimates for diverse populations.
 The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
-Here, by carefully using information on self-reported ancestry, country of birth and genetic similarity, we define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-This includes five groups with genetic ancestries from Africa, eight from Europe, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:counts}).
+Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 20 distinct ancestry groups from the UK Biobank to be used as global reference populations.
+This includes six groups with genetic ancestries from Europe, six from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
 The detailed procedure used to define these groups is presented in the Supplementary Materials.
-We provide [TODO FINISH \#VARIANTS].
+We then provide reference allele frequencies for 5,816,590 genetic variants across these 20 diverse ancestry groups.
 
-\begin{table}[h]
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Fri Oct 22 15:40:44 2021
+\begin{table}[!htb]
 	\centering
-	\begin{tabular}{|l|c||l|c||l|c|}
+	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c|c|c||c|c|c|c|c|c|c|c|}
 		\hline
-		Ancestry group & N & Ancestry group & N & Ancestry group & N \\
+		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & AMR & Caribbean & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
 		\hline
-		Japan & 240 (+ 104) & Asia (East) & 504 & Philippines & 304 \\
-		Caribbean & 660 & Africa (West) & 655 & Africa (South) & 362 \\
-		Africa (East 1) & 193 & Africa (East 2) & 134 & Middle East & 509 \\
-		Pakistan & 388 & Sri Lanka & 365 & Bangladesh & 223 (+ 86) \\
-		Europe (North West) & 54556 & Finland & 143 (+ 99) & Scandinavia & 342 \\
-		Europe (North East) & 259 & Europe (South East) & 344 & Europe (South West) & 605 \\
-		Europe (Central) & 559 & Ashkenazi & 1975 & South America & 560 \\
+		Africa (West) & 735 &  &  &  &  & 34.4 & 1.2 & 1.7 & 22.6 & 1.3 & 72.5 & 0.2 & 0.2 & 2 & 0.6 & 3.4 & 0.3 & 0.5 & 2.2 \\ 
+		Africa (Central) & 276 &  &  &  &  & 4.4 & 0.3 & 0.5 & 6.9 & 0.1 & 15.1 & 0.1 &  & 0.7 & 0.7 & 1.1 & 0.1 & 0.2 & 2.3 \\ 
+		Africa (South) & 449 &  &  &  &  & 61.1 & 0.2 & 0.2 & 3 & 0.5 & 3.6 &  & 0.1 & 0.4 & 0.8 & 0.5 & 0.1 &  & 2.3 \\ 
+		Africa (East 2) & 279 &  &  &  &  &  &  &  & 0.7 &  & 1 &  &  & 0.3 & 0.1 & 0.3 &  &  & 1.1 \\ 
+		\hline
+		Africa (East 1) & 276 &  &  &  & 11.4 &  &  &  &  &  &  &  &  &  & 0.3 & 0.1 &  &  & 1.1 \\ 
+		Africa (North) & 268 &  &  &  & 22.1 &  &  &  & 1.5 & 1.1 &  &  &  &  &  &  &  &  &  \\ 
+		Middle East & 523 &  &  &  & 63.5 &  &  & 0.9 &  &  &  &  &  & 2.8 & 0.1 & 0.1 &  & 0.7 & 1 \\ 
+		\hline
+		Ashkenazi & 1975 &  &  &  & 1.9 &  & 0.9 & 5 & 0.4 &  & 0.1 & 0.3 & 2.3 & 1.3 & 1.3 & 0.6 & 0.8 & 2.3 & 0.8 \\ 
+		Italy & 345 &  &  &  &  &  & 2.7 & 6 & 2.3 & 0.9 & 0.1 & 4 & 4.3 & 3.4 & 4.5 & 1 & 4.1 & 6.4 & 3.1 \\ 
+		Europe (East) & 1223 &  & 7.5 &  &  &  & 3 & 9.8 & 0.5 &  &  & 11.9 & 12.7 & 13.2 & 15 & 15 & 12.9 & 11.8 & 10.2 \\ 
+		Finland & 242 &  & 85.8 &  &  &  &  & 1.3 & 0.5 & 0.2 &  & 9 & 12.2 & 6.1 & 8 & 13 & 10.4 & 4.7 & 2.4 \\ 
+		Europe (North West) & 4416 &  & 6 &  &  &  & 81.2 & 53.8 & 8.4 &  & 6.2 & 62.8 & 58.4 & 47.5 & 50.6 & 56.5 & 63.5 & 62.3 & 37.8 \\ 
+		Europe (South West) & 603 &  &  &  &  &  & 7.7 & 8.4 & 8.2 & 6.7 & 1 & 7 & 7.1 & 4.3 & 5 & 3.1 & 6.4 & 7.5 & 4.3 \\ 
+		\hline
+		South America & 473 &  & 0.3 & 94.7 &  &  & 0.2 & 5.1 & 27.7 & 89.0 & 0.3 & 1.1 & 0.8 & 1.6 & 0.8 & 2.1 & 0.6 & 0.3 & 0.3 \\ 
+		\hline
+		Bangladesh & 309 &  &  &  &  &  & 0.6 & 0.3 & 1.3 &  &  & 1.2 & 1 & 2.7 & 2.7 & 0.9 &  &  & 1.3 \\ 
+		Pakistan & 400 &  &  &  &  &  & 1.4 &  & 0.1 &  &  & 0.6 &  & 6.6 & 3.1 & 1.1 &  &  & 0.3 \\ 
+		Sri Lanka & 372 &  &  &  & 1.1 &  & 0.3 & 0.1 & 0.7 &  &  & 0.9 & 0.4 & 2.2 & 1.6 & 0.3 &  &  & 0.7 \\ 
+		\hline
+		Philippines & 295 &  &  &  &  &  & 0.1 & 1.4 & 3.9 &  &  &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.5 \\ 
+		Asia (East) & 961 & 3.9 &  &  &  &  & 0.3 & 3.8 & 2.6 &  &  & 0.6 & 0.3 & 2.5 & 1.8 & 0.3 &  & 3.1 & 2.4 \\ 
+		Japan & 344 & 96.1 & 0.5 & 5.3 &  &  &  & 1.9 & 8.7 &  &  & 0.3 & 0.3 & 2.2 & 3.1 & 0.5 & 0.6 &  & 26 \\ 
 		\hline
 	\end{tabular}
-	\caption{Number of individuals (N) in each of the 21 ancestry groups we define in the UK Biobank. Numbers in parentheses represent the added individuals from the 1000 Genomes data.}
-	\label{tab:counts}
+	\caption{Reference population with their size (N) and corresponding ancestry proportions in other GWAS datasets.
+Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, admixed Americans in the 1000 Genomes data (AMR, \cite{10002015global}), Caribbean ancestry group in the UKBB (Supplementary Materials), breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
+Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
+\label{tab:table1}}
 \end{table}
 
-Then, we propose to estimate ancestry proportions from a cohort based on its allele frequencies and the reference set of allele frequencies described previously.
-We propose to find the mixture proportions $\alpha_k$ which minimize the following problem:
+Then, as a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only.
+We propose to find the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimize the following problem:
 \(\sum_{j=1}^M \left(w_j  f_j^{(0)} - \sum_{k=1}^K \alpha_k w_j f_j^{(k)} \right)^2 ,\)
 where $M$ is the number of variants, $K$ the number of reference populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the cohort of interest.
-By further constraining all $w_j$ to be positive and sum to 1 (convex combination) in the optimization, we can interpret these coefficients as ancestry proportions.
-This is similar to Summix \cite[]{arriaga2021summix}, although we introduce weights $w_j = 1 / \sqrt{f_j^{(0)} (1 - f_j^{(0)})}$ to account for more variability in more common variants.
-This optimization problem is a standard quadratic programming problem that can be solved using R package quadprog \cite[]{turlach2013quadprog}, and that we implement in function [TODO] in our R package bigsnpr \cite[]{prive2017efficient}.
+This is similar to Summix \cite[]{arriaga2021summix}, except for two differences.
+First, we introduce weights $w_j = 1 / \max\left(\sqrt{f_j^{(0)} (1 - f_j^{(0)})}, ~0.1\right)$ to account for more variability in more common variants.
+Second, and most importantly, we use the new reference set of allele frequencies we present here (instead of the five continental 1000 Genomes populations) for measuring genetic diversity with more precision.
+This optimization problem is a standard quadratic programming problem that can be solved using R package quadprog \cite[]{turlach2013quadprog}, and that we implement in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}.
 
-%% Around 320 words already %%
+We download several GWAS summary statistics which reports the allele frequencies and apply this method to both validate our new reference set we provide and show one of its potential use.
+We first apply function \texttt{snp\_ancestry\_summary} to homogeneous samples as an empirically validation.
+When applied to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, our function using the new reference set we provide clearly estimate ancestry proportions matching expectations.
+When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries, whereas our estimates are of 27.7\% South American, 20.2\% European, 34.8\% African, 2.1\% South Asian, 11.3\% East Asian, 3.9\% Filipino.
+GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented.
 
-[TODO: CONCLU] Using allele frequencies derived from these groups, we are now able to effectively measure diversity from summary statistics of a genetic dataset.
+Here we provide an unprecedentedly large and diverse reference set of allele frequencies using the UK Biobank.
+We then show how to effectively measure diversity from GWAS summary statistics using this new set of reference allele frequencies.
 Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
-
-\begin{figure}[!htpb]
-	%\centerline{\includegraphics[width=0.7\textwidth]{timings.pdf}}
-	\caption{Computation times of \texttt{pcadapt} as a function of the number of PCs used.
-Black points represent timings with pcadapt v4, and the dashed blue line represents the timing with v3, which is independent of the number of PC used.
-\label{fig:timings}}
-\end{figure}
+Our work has limitations though.
+First, although we managed to define ancestry groups with more than 240 individuals, it seems that they are not enough to recover for our method to recover 100\% of one ancestry (e.g.\ the case for the FinnGen and Biobank Japan cohorts which are likely very homogenous).
+In this case, we also capture a small proportion from nearby ancestry groups.
+Second, it is not fully clear whether we can capture any ancestry, e.g. Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
+Third, with the 20 ancestry groups we defined, we think we are capturing a large amount of the diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in the genetic ancestry  in both South Asia and South America.
+Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing some quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
+Anyway, we envision that the new set of reference allele frequencies we provide will have many applications.
 

### 2021-10-24 paper-ancestry-prop.tex
save

@@
 
-[TODO]
+The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
+The set of reference allele frequencies for 5,816,590 genetic variants across 20 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31163785}.
+All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 
+We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
 
@@
 
-%\clearpage
+\clearpage
 
@@
 
-\clearpage
-%\section*{Supplementary Materials}
-%
-%\renewcommand{\thefigure}{S\arabic{figure}}
-%\setcounter{figure}{0}
-%\renewcommand{\thetable}{S\arabic{table}}
-%\setcounter{table}{0}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2021-10-27 paper-ancestry-prop-supp.tex
save
add admixture and restrict to 17 groups

@@
 Finally, we keep only individuals born in South Africa with a log-distance to the UK larger than 6 (individuals of African ancestry). 
-After applying these filters, there are still many individuals born outside of the UK, including e.g.\ 3352 individuals from India, 1100 from Germany, 1036 from Nigeria, and 76 countries with at least 50 individuals.
+After applying these filters, there are still many individuals born outside of the UK and Ireland, including e.g.\ 3352 individuals from India, 1100 from Germany, 1036 from Nigeria, and 76 countries with at least 50 individuals.
 
-Then, to define a homogeneous group to be used as a reference population, we choose a country, compute the robust PC center of individuals born in this country using the geometric median \cite[]{prive2021high}, and choose to include in this group all individuals within a chosen log-distance to the center.
+Then, to define a homogeneous group to be used as a reference population, we choose a country, compute the robust PC center of individuals born in this country (using the geometric median, \cite{prive2021high}), and choose to include in this group all individuals within a chosen log-distance to the center.
 The distance threshold is chosen based on the countries of birth included.
@@
 We manually try different country centers and distance thresholds to cover as much of the worldwide populations as possible, without having overlapping groups.
-We also include two other groups: one ``Ashkenazi'' group based on a PC center we defined in \cite{prive2021high}, and a ``South America'' group based on individuals with $\text{PC6} > 50$ and $\text{PC8} < -30$ (clearly separated from the rest of individuals from the UKBB).
-Finally, we investigate individuals that are far away from any of the groups defined previously, and add a last group based on a center from India.
+We also include two other groups: one ``Ashkenazi'' group based on a PC center we defined in \cite{prive2021high}, and a ``South America'' group based on individuals with $\text{PC6} > 50$ and $\text{PC8} < -30$ (clearly separated from the rest of individuals from the UK Biobank).
+Finally, we investigate individuals that are far away from any of the groups defined previously, and add a last group based on a center from remaining individuals born in India.
 
@@
 
-\subsection{Designing the final sets of allele frequencies}
+\subsection{Designing the final set of variants}
 
@@
 Based on the remaining 5,840,630 variants, we compute the overall fixation indices $F_{ST}$ between the UK Biobank groups and the 1KG populations (See the description of the 1KG populations at \url{https://www.coriell.org/1/NHGRI/Collections/1000-Genomes-Collections/1000-Genomes-Project}).
-This enable us to check that the group we defined appropriate capture the ancestries they should.
-We report all $F_{ST}$ in a CSV file, and show only the lowest for each of the 26 ancestry groups we define here:
+We report all $F_{ST}$ at \url{https://github.com/privefl/freq-ancestry/blob/main/all_fst.csv}, and show only the lowest for each of the 26 ancestry groups we define here:
 \begin{itemize}
@@
 For each of the 1KG populations, we identify the closest ancestry group among those we defined, compute the $F_{ST}$ for all variants and scale them by dividing by the overall $F_{ST}$ with this group.
-For each variant, we then sum these scaled $F_{ST}$ (26 different ones corresponding to the UK Biobank ancestry groups), and identify and remove 18,344 variants with a large score.
+For each variant, we then sum these scaled $F_{ST}$ (one for each 1KG population), and identify and remove 24,040 variants with a large score.
 This enable us to identify variants with possible errors in the UK Biobank.
 
-Finally, for the Finland, Bangladesh and Japan ancestry groups, we compute a weighted average of allele frequencies with the ones from the 1KG data (FIN, BEB and JPT) to increase the small sample size of these groups.
-Since some ancestry groups are genetically very close to each other (very low $F_{ST}$), we merge them by averaging their allele frequencies.
-Thus we merge the ``Irish'', ``United Kingdom'' and ``Scandinavia'' groups  to form a new Northwestern European group.
-We also merge the groups from Central, North-East and South-East Europe to form a new Eastern European group.
-We remove the ``India'' group since the $F_{ST}$ with other populations seem to place this group north of India, instead of in the middle of India, as we sought.
-We also remove the ``Caribbean'' group since it is an admixed population (estimated by our method to be of 92.2\% African, 6.2\% Northwestern European and 1\% Southwestern European ancestries, see Table 1 of main text).
+\subsection{Designing the final set of reference populations}
 
-This results in the set of 5,816,590 variants for 20 populations, which we provide for download, and that we use to estimate ancestry proportions in GWAS summary statistics.
+The 26 ancestry groups defined previously comprise a total of 15,645 individuals from the UK Biobank.
+We combine them with 2490 individuals from the 26 populations of the 1KG data, and use a pruned set of 252,893 high-quality genotyped variants to run an admixture analysis (K = 15) using function \texttt{snmf} from R package LEA \cite[]{frichot2014fast,gain2020lea}.
+Ancestry proportions for the 52 groups are represented in Figure \ref{fig:admixture}.
+We can see that we have defined very homogeneous groups; however based on these results and previous $F_{ST}$, we decide to slightly modify these groups.
+First, for the Finland, Bangladesh and Japan ancestry groups, we compute a weighted average of allele frequencies with the ones from the 1KG data (FIN, BEB and JPT) to increase the small sample size of these groups.
+We do the same for ``South America'' and ``PEL'' in order to capture slightly more of what is probably an Amerindian ancestral component (in blue).
+Second, we merge the ``Irish'' and ``United Kingdom'' groups as a new ``British-Irish Isles'' group.
+Third, we remove some groups that could be retrieved from combining two or three other groups: ``Africa (East 2)'', ``Caribbean'', ``Africa (Central)'', ``Scandinavia'', ``Italy'', ``Europe (South East)'', ``Europe (Central)'', and ``India''.
+
+\begin{figure}[hp]
+	\centerline{\includegraphics[width=0.9\textwidth]{admixture}}
+	\caption{Admixture coefficients (for K = 15 ancestral populations) for the individuals from the 26 ancestry groups we defined previously from the UK Biobank, as well as the 26 1KG populations.
+\label{fig:admixture}}
+\end{figure}
+
+This results in the set of 5,816,590 variants for 17 ancestry groups, which we provide for download and use to estimate ancestry proportions in GWAS summary statistics.
 

### 2021-10-27 paper-ancestry-prop.tex
add admixture and restrict to 17 groups

@@
 
-\title{Using the UK Biobank as a global reference: application to measuring ancestry diversity from GWAS summary statistics}
+\title{Using the UK Biobank as a global reference of worldwide populations: application to measuring ancestry diversity from GWAS summary statistics}
 \author{Florian Priv\'e$^{\text{1}}$}
@@
 	The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
-	Within this dataset, we carefully define 20 distinct ancestry groups from all four corners of the world. 
+	Within this dataset, we carefully define 17 distinct ancestry groups from all four corners of the world. 
 	Using allele frequencies derived from these global reference groups, we are now able to effectively measure diversity from summary statistics of any genetic dataset.
@@
 Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project, the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
-However, these datasets do not contain many individuals per population and therefore are large enough to provide accurate allele frequency estimates for diverse populations.
+However, these datasets do not contain many individuals per population and therefore are not large enough to provide accurate allele frequency estimates for diverse worldwide populations.
 The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
-Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 20 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-This includes six groups with genetic ancestries from Europe, six from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
+Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 17 distinct ancestry groups from the UK Biobank to be used as global reference populations.
+This includes five groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
 The detailed procedure used to define these groups is presented in the Supplementary Materials.
-We then provide reference allele frequencies for 5,816,590 genetic variants across these 20 diverse ancestry groups.
+We then provide reference allele frequencies for 5,816,590 genetic variants across these 17 diverse ancestry groups.
 
@@
 	\end{tabular}
-	\caption{Reference population with their size (N) and corresponding ancestry proportions in other GWAS datasets.
-Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, admixed Americans in the 1000 Genomes data (AMR, \cite{10002015global}), Caribbean ancestry group in the UKBB (Supplementary Materials), breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
+	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets.
+Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
 Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
@@
 
-Then, as a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only.
+Then, as a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
 We propose to find the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimize the following problem:
@@
 
-We download several GWAS summary statistics which reports the allele frequencies and apply this method to both validate our new reference set we provide and show one of its potential use.
-We first apply function \texttt{snp\_ancestry\_summary} to homogeneous samples as an empirically validation.
-When applied to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, our function using the new reference set we provide clearly estimate ancestry proportions matching expectations.
-When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries, whereas our estimates are of 27.7\% South American, 20.2\% European, 34.8\% African, 2.1\% South Asian, 11.3\% East Asian, 3.9\% Filipino.
-GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented.
+We download several GWAS summary statistics which reports allele frequencies and apply this method to both validate our new reference set we provide and show one of its potential use.
+We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirically validation.
+When applying this function (using the new reference set we provide) to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions estimates obtained match expectations (Table \ref{tab:table1}).
+When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 27.7\% South American, 20.2\% European, 34.8\% African, 2.1\% South Asian, 11.3\% East Asian, and 3.9\% Filipino [TODO: CHANGE].
+GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented (Table \ref{tab:table1}).
 
@@
 Our work has limitations though. 
-First, although we managed to define ancestry groups with more than 240 individuals, it seems that they are not enough to recover for our method to recover 100\% of one ancestry (e.g.\ the case for the FinnGen and Biobank Japan cohorts which are likely very homogenous).
+First, although we managed to define ancestry groups with more than 240 individuals, it seems that these groups are still not large enough for the ancestry deconvolution method to recover 100\% of one ancestry (e.g.\ the case for the FinnGen and Biobank Japan cohorts which are likely very homogenous).
 In this case, we also capture a small proportion from nearby ancestry groups.
-Second, it is not fully clear whether we can capture any ancestry, e.g. Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
-Third, with the 20 ancestry groups we defined, we think we are capturing a large amount of the diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in the genetic ancestry  in both South Asia and South America.
-Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing some quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
-Anyway, we envision that the new set of reference allele frequencies we provide will have many applications.
+Second, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
+Third, with the 17 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in both South Asia and South America, as well as in some isolated populations.
+Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
+Anyway, despite these limitations, we envision that the new set of reference allele frequencies we provide will have many useful applications.
 
@@
 The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-The set of reference allele frequencies for 5,816,590 genetic variants across 20 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31163785}.
+The set of reference allele frequencies for 5,816,590 genetic variants across 17 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31163785} [TODO: REUPLOAD AND VERIFY].
 All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 

### 2021-10-27 paper-ancestry-prop.tex
reexport code

@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 
-% latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Fri Oct 22 15:40:44 2021
-\begin{table}[!htb]
-	\centering
-	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c|c|c||c|c|c|c|c|c|c|c|}
-		\hline
-		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & AMR & Caribbean & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
-		\hline
-		Africa (West) & 735 &  &  &  &  & 34.4 & 1.2 & 1.7 & 22.6 & 1.3 & 72.5 & 0.2 & 0.2 & 2 & 0.6 & 3.4 & 0.3 & 0.5 & 2.2 \\ 
-		Africa (Central) & 276 &  &  &  &  & 4.4 & 0.3 & 0.5 & 6.9 & 0.1 & 15.1 & 0.1 &  & 0.7 & 0.7 & 1.1 & 0.1 & 0.2 & 2.3 \\ 
-		Africa (South) & 449 &  &  &  &  & 61.1 & 0.2 & 0.2 & 3 & 0.5 & 3.6 &  & 0.1 & 0.4 & 0.8 & 0.5 & 0.1 &  & 2.3 \\ 
-		Africa (East 2) & 279 &  &  &  &  &  &  &  & 0.7 &  & 1 &  &  & 0.3 & 0.1 & 0.3 &  &  & 1.1 \\ 
-		\hline
-		Africa (East 1) & 276 &  &  &  & 11.4 &  &  &  &  &  &  &  &  &  & 0.3 & 0.1 &  &  & 1.1 \\ 
-		Africa (North) & 268 &  &  &  & 22.1 &  &  &  & 1.5 & 1.1 &  &  &  &  &  &  &  &  &  \\ 
-		Middle East & 523 &  &  &  & 63.5 &  &  & 0.9 &  &  &  &  &  & 2.8 & 0.1 & 0.1 &  & 0.7 & 1 \\ 
-		\hline
-		Ashkenazi & 1975 &  &  &  & 1.9 &  & 0.9 & 5 & 0.4 &  & 0.1 & 0.3 & 2.3 & 1.3 & 1.3 & 0.6 & 0.8 & 2.3 & 0.8 \\ 
-		Italy & 345 &  &  &  &  &  & 2.7 & 6 & 2.3 & 0.9 & 0.1 & 4 & 4.3 & 3.4 & 4.5 & 1 & 4.1 & 6.4 & 3.1 \\ 
-		Europe (East) & 1223 &  & 7.5 &  &  &  & 3 & 9.8 & 0.5 &  &  & 11.9 & 12.7 & 13.2 & 15 & 15 & 12.9 & 11.8 & 10.2 \\ 
-		Finland & 242 &  & 85.8 &  &  &  &  & 1.3 & 0.5 & 0.2 &  & 9 & 12.2 & 6.1 & 8 & 13 & 10.4 & 4.7 & 2.4 \\ 
-		Europe (North West) & 4416 &  & 6 &  &  &  & 81.2 & 53.8 & 8.4 &  & 6.2 & 62.8 & 58.4 & 47.5 & 50.6 & 56.5 & 63.5 & 62.3 & 37.8 \\ 
-		Europe (South West) & 603 &  &  &  &  &  & 7.7 & 8.4 & 8.2 & 6.7 & 1 & 7 & 7.1 & 4.3 & 5 & 3.1 & 6.4 & 7.5 & 4.3 \\ 
-		\hline
-		South America & 473 &  & 0.3 & 94.7 &  &  & 0.2 & 5.1 & 27.7 & 89.0 & 0.3 & 1.1 & 0.8 & 1.6 & 0.8 & 2.1 & 0.6 & 0.3 & 0.3 \\ 
-		\hline
-		Bangladesh & 309 &  &  &  &  &  & 0.6 & 0.3 & 1.3 &  &  & 1.2 & 1 & 2.7 & 2.7 & 0.9 &  &  & 1.3 \\ 
-		Pakistan & 400 &  &  &  &  &  & 1.4 &  & 0.1 &  &  & 0.6 &  & 6.6 & 3.1 & 1.1 &  &  & 0.3 \\ 
-		Sri Lanka & 372 &  &  &  & 1.1 &  & 0.3 & 0.1 & 0.7 &  &  & 0.9 & 0.4 & 2.2 & 1.6 & 0.3 &  &  & 0.7 \\ 
-		\hline
-		Philippines & 295 &  &  &  &  &  & 0.1 & 1.4 & 3.9 &  &  &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.5 \\ 
-		Asia (East) & 961 & 3.9 &  &  &  &  & 0.3 & 3.8 & 2.6 &  &  & 0.6 & 0.3 & 2.5 & 1.8 & 0.3 &  & 3.1 & 2.4 \\ 
-		Japan & 344 & 96.1 & 0.5 & 5.3 &  &  &  & 1.9 & 8.7 &  &  & 0.3 & 0.3 & 2.2 & 3.1 & 0.5 & 0.6 &  & 26 \\ 
-		\hline
-	\end{tabular}
-	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets.
-Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
-Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
-\label{tab:table1}}
-\end{table}
-
 Then, as a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
@@
 When applying this function (using the new reference set we provide) to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions estimates obtained match expectations (Table \ref{tab:table1}).
-When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 27.7\% South American, 20.2\% European, 34.8\% African, 2.1\% South Asian, 11.3\% East Asian, and 3.9\% Filipino [TODO: CHANGE].
+When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 24.9\% South American, 21.9\% European, 35.4\% African, 2.0\% South Asian, 11.3\% East Asian, 3.9\% Filipino, and 0.5\% Middle East.
 GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented (Table \ref{tab:table1}).
+
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Wed Oct 27 13:40:20 2021
+\begin{table}[htb]
+	\centering
+	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c||c|c|c|c|c|c|c|c|}
+		\hline
+		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
+		\hline
+		Africa (West) & 735 &  &  &  &  & 36.7 & 1.4 & 1.9 & 26.4 & 0.2 & 0.2 & 2.4 & 1 & 4 & 0.3 & 0.6 & 3.6 \\ 
+		Africa (South) & 449 &  &  &  &  & 63.3 & 0.3 & 0.4 & 6.8 &  & 0.1 & 0.9 & 1.1 & 1.2 & 0.1 &  & 4 \\ 
+		Africa (East) & 276 &  &  &  & 11.4 &  &  &  & 0.2 &  &  &  & 0.2 & 0.1 &  &  & 1.4 \\ 
+		Africa (North) & 268 &  &  &  & 22.1 &  &  & 0.4 & 2 &  &  &  &  &  &  & 0.3 &  \\ 
+		\hline
+		Middle East & 523 &  &  &  & 63.5 &  & 1.3 & 3.4 & 0.5 & 2 & 2.2 & 5.2 & 2.9 & 1.5 & 1.6 & 3.2 & 2.8 \\ 
+		\hline
+		Ashkenazi & 1975 &  &  &  & 1.9 &  & 1.4 & 6 & 0.9 & 1.3 & 3.3 & 2.2 & 2.3 & 1.3 & 2 & 3.6 & 1.5 \\ 
+		Europe (North East) & 320 &  & 8.9 &  &  &  & 3.8 & 7.7 & 0.7 & 9.6 & 10 & 10 & 11 & 11.7 & 10.1 & 8.9 & 7.5 \\ 
+		Finland & 143 (+ 99) &  & 85.2 &  &  &  & 2.7 & 3.5 & 0.7 & 11.9 & 14.8 & 8.2 & 10.5 & 15.4 & 13.1 & 7.1 & 4.3 \\ 
+		British-Irish Isles & 4000 &  & 5.1 &  &  &  & 79.1 & 53 & 8.9 & 59.8 & 56.1 & 46.6 & 49.2 & 54.8 & 61.7 & 62 & 36.6 \\ 
+		Europe (South West) & 603 &  &  &  &  &  & 7.7 & 12.2 & 10.7 & 11.2 & 11.2 & 7.5 & 9.5 & 5.3 & 10.1 & 11.2 & 7.3 \\ 
+		\hline
+		South America & 473 (+ 84) &  & 0.4 & 99.1 &  &  & 0.1 & 4.5 & 24.9 & 0.9 & 0.6 & 1.4 & 0.7 & 1.8 & 0.4 & 0.1 & 0.2 \\ 
+		\hline
+		Japan & 240 (+ 104) & 96.1 & 0.4 & 0.9 &  &  &  & 1.8 & 8.8 & 0.2 & 0.2 & 2.1 & 3 & 0.4 & 0.4 &  & 25.9 \\ 
+		Asia (East) & 961 & 3.9 &  &  &  &  & 0.2 & 3.7 & 2.5 & 0.6 & 0.3 & 2.5 & 1.8 & 0.2 &  & 3 & 2.3 \\ 
+		Philippines & 295 &  &  &  &  &  & 0.1 & 1.4 & 3.9 &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.5 \\ 
+		\hline
+		Bangladesh & 223 (+ 86) &  &  &  &  &  & 0.6 &  & 1.2 & 1.1 & 0.8 & 2.8 & 2.8 & 0.9 &  &  & 1.4 \\ 
+		Sri Lanka & 372 &  &  &  & 1.1 &  & 0.4 &  & 0.8 & 0.8 & 0.2 & 2.2 & 1.6 & 0.4 &  &  & 0.7 \\ 
+		Pakistan & 400 &  &  &  &  &  & 0.8 &  &  & 0.4 &  & 6 & 2.5 & 0.7 &  &  &  \\ 
+		\hline
+	\end{tabular}
+	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets.
+Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
+Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
+\label{tab:table1}}
+\end{table}
 
@@
 Second, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
-Third, with the 17 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in both South Asia and South America, as well as in some isolated populations.
+Third, with the 17 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in both South Asia and South America.
 Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
@@
 The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-The set of reference allele frequencies for 5,816,590 genetic variants across 17 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31163785} [TODO: REUPLOAD AND VERIFY].
+The set of reference allele frequencies for 5,816,590 genetic variants across 17 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31218445}.
 All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 

### 2021-10-27 code/1-prepare-ukbb-groups.R

@@ -81,8 +81,6 @@ qplot(log(dist_to_UK[country %in% POP]), geom = "density",
   geom_vline(xintercept = 4, linetype = 3) +
   # geom_vline(xintercept = c(2.5, 5), linetype = 3) +
   theme(legend.position = c(0.2, 0.7))
-# ggsave("figures/noeur-dist-uk.pdf", width = 8, height = 6, scale = 0.8)
-# ggsave("figures/eur-dist-uk.pdf", width = 8, height = 6, scale = 0.8)
 
 # Probably European:
 country[log(dist_to_UK) < 4 & continent != "Europe"] <- NA

### 2021-10-27 code/2-run-admixture.R

@@ -1,0 +1,119 @@
+#### Prepare variants to use ####
+
+snp_qc <- bigreadr::fread2(runonce::download_file(
+  "https://biobank.ctsu.ox.ac.uk/crystal/crystal/auxdata/ukb_snp_qc.txt",
+  dir = "tmp-data"
+))
+
+af_qc <- bigreadr::fread2(runonce::download_file(
+  "http://kunertgraf.com/data/files/snp_maf_comparison.csv",
+  dir = "tmp-data"
+))
+
+
+library(dplyr)
+snp_qc$qc_all <- rowMeans(select(snp_qc, ends_with("_qc")))
+snp_qc2 <- snp_qc %>%
+  select(-ends_with("_qc"), -ends_with("_loading")) %>%
+  left_join(af_qc, by = c("rs_id" = "rsid"))
+
+plot(as.factor(snp_qc2$oob), snp_qc2$qc_all)
+
+snp_qc_final <- snp_qc %>%
+  select(-2, -3, -ends_with("_qc"), -ends_with("_loading")) %>%
+  as_tibble() %>%
+  filter(array == 2, qc_all == 1) %>%
+  anti_join(filter(af_qc, oob == 1), by = c("rs_id" = "rsid")) %>%
+  select(2:5)
+
+library(doParallel)
+registerDoParallel(cl <- makeCluster(12))
+system.time({
+  list_snp_id <- foreach(chr = 1:22, .packages = "dplyr") %dopar% {
+    paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt") %>%
+      bigreadr::fread2(showProgress = FALSE) %>%
+      filter(V6 > 0.01, V8 == 1) %>%   ## MAF > 1% & INFO = 1
+      dplyr::semi_join(
+        filter(snp_qc_final, chromosome == !!chr),
+        by = c("V3" = "position", "V4" = "allele1_ref", "V5" = "allele2_alt")
+      ) %>%
+      with(paste(chr, V3, V4, V5, sep = "_"))
+  }
+}) # 40 sec
+stopCluster(cl)
+
+sum(lengths(list_snp_id))  # 586,534
+
+obj.bed <- bed(download_1000G("tmp-data"))
+list_snp_id_1kg <- with(obj.bed$map, paste(chromosome, physical.pos, allele2, allele1, sep = "_"))
+
+list_snp_id2 <- lapply(list_snp_id, intersect, y = list_snp_id_1kg)
+sum(lengths(list_snp_id2))  # 470,878
[…]

### 2021-10-27 code/3-prepare-ukbb-freq.R

@@ -15,7 +15,7 @@ grid <- purrr::map_dfr(1:22, function(chr) {
 })
 nrow(grid) # 257 parts
 
-all_eid <- readRDS("tmp-data/info-UKBB.rds")$eid
+all_eid <- readRDS("tmp-data/info-UKBB2.rds")$eid
 eid_sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
 all_ind <- lapply(readRDS("data/list_ind_pop.rds"), function(ind) {
   match(all_eid[ind], eid_sample)

### 2021-10-27 code/4-filter-merge-freq.R

@@ -56,7 +56,7 @@ foreach(i = 2:npop, .export = "all_fst") %:%
 all_fst <- all_fst[]
 colnames(all_fst) <- rownames(all_fst) <- names(all_N)
 # write.table(all_fst, "all_fst.csv", sep = ",", quote = FALSE)
-# read.csv("all_fst.csv", check.names = FALSE)
+# all_fst <- as.matrix(read.csv("all_fst.csv", check.names = FALSE))
 
 print_fst <- function(pop, nb = 7) {
   smallest_fst <- sort(all_fst[, pop])[1:nb + 1L]
@@ -119,13 +119,16 @@ X <- as.data.frame(X_ukbb[ind, ])
 X[, "Finland"] <- (143 * X[, "Finland"] + 99 * X_1kg[ind, "FIN"]) / (143 + 99)
 X[, "Bangladesh"] <- (223 * X[, "Bangladesh"] + 86 * X_1kg[ind, "BEB"]) / (223 + 86)
 X[, "Japan"] <- (240 * X[, "Japan"] + 104 * X_1kg[ind, "JPT"]) / (240 + 104)
+X[, "South America"] <- (473 * X[, "South America"] + 84 * X_1kg[ind, "PEL"]) / (473 + 84)
 # Merge very close ancestry groups (based on Fst above)
-X[, "Europe (North West)"] <- (X$Ireland + X$`United Kingdom` + X$Scandinavia) / 3
-X[, "Europe (East)"] <- (X$`Europe (North East)` + X$`Europe (Central)` + X$`Europe (South East)`) / 3
+X[, "British-Irish Isles"] <- (X$Ireland + X$`United Kingdom`) / 2
 # Remove India (not totally expected Fst) and Caribbean (too close to West Africa and admixed)
-X2 <- dplyr::select(X, -c(Ireland, `United Kingdom`, Scandinavia, `Europe (North East)`,
-                          `Europe (Central)`, `Europe (South East)`, India, Caribbean))
+X2 <- dplyr::select(X, -c(Ireland, `United Kingdom`, `Africa (East 2)`, Caribbean, `Africa (Central)`,
+                          Scandinavia, Italy,`Europe (South East)`, `Europe (Central)`, India),
+                    `Africa (East)` = `Africa (East 1)`)
 
-ord <- hclust(as.dist(1 - cor(X2)^2), method = "ward.D2")$order
[…]

### 2021-10-27 code/5-assess-ancestry-in-sumstats.R

@@ -32,33 +32,6 @@ str(res <- snp_match_ancestry(sumstats))
 saveRDS(res, "res_ancestry/UKBB.rds")
 
 
-## Caribbean defined previously
-
-sumstats <- cbind.data.frame(
-  readRDS("tmp-data/info_af_1kg.rds")[1:5],
-  readRDS("tmp-data/info_af_ukbb.rds")$Caribbean
-) %>%
-  filter(info > 0.4)
-
-str(res <- snp_match_ancestry(sumstats))
-saveRDS(res, "res_ancestry/caribbean.rds")
-
-
-
-## AMR in 1KG
-
-obj.bed <- bed(download_1000G("tmp-data"))
-fam2 <- bigreadr::fread2(paste0(obj.bed$prefix, ".fam2"))
-ind_row <- which(fam2$`Super Population` == "AMR")
-
-af <- bed_MAF(obj.bed, ind_row, ncores = nb_cores())$af
-sumstats <- dplyr::transmute(obj.bed$map, chr = chromosome, pos = physical.pos,
-                             a1 = allele1, a0 = allele2, freq = af)
-
-str(res <- snp_match_ancestry(sumstats))
-saveRDS(res, "res_ancestry/AMR.rds")
-
-
 ## T2D from Biobank Japan
 
 # zip <- runonce::download_file(
@@ -234,7 +207,7 @@ sumstats <- fread2(
   col.names = c("chr", "pos", "a0", "a1", "freq"),
 )
 
-str(res <- snp_match_ancestry(sumstats))  # only 0.95 correlation
+str(res <- snp_match_ancestry(sumstats))
 saveRDS(res, "res_ancestry/height-peru.rds")
 
 
@@ -310,7 +283,22 @@ all_coef <- setNames(purrr::map_dfc(files, ~ readRDS(.x)$coef),
 
 library(dplyr)
 pop_names <- names(readRDS("res_ancestry/UKBB.rds")$coef)
-pop_N <- lengths(readRDS("data/list_ind_pop.rds"))
+pop_N <- c(lengths(readRDS("data/list_ind_pop.rds")), "British-Irish Isles" = 4000)
+pop_N["Africa (East)"] <- pop_N["Africa (East 1)"]
 csv <- bind_cols(Population = pop_names, N = pop_N[pop_names], all_coef) %>%
   print(n = Inf, width = Inf) %>%
   bigreadr::fwrite2("all_prop.csv")
+
[…]

### 2021-11-09 paper-ancestry-prop.tex
save

@@
 		\hline
-		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
+		Ancestry group & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
 		\hline

### 2021-11-09 pres/group-meeting.Rmd
save

---
title: "Group meeting"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 70)
knitr::opts_chunk$set(fig.align = 'center', dev = "svg", out.width = "70%",
                      echo = FALSE, comment = "", fig.width = 5, global.par = TRUE)
```

class: title-slide center middle inverse

<br>

# Global ancestry proportions and frequencies from the UK Biobank

<br>

## Florian Privé

---

## Goal

<br>

#### Infer ancestry proportions using allele frequencies only

---

## Main challenges

<br>

- Need some (precise) frequencies from many global populations

--

- Need a method to infer ancestry proportions.    
    We can minimize the following problem, subject to $w \ge 0$ and $\sum_k w_k = 1$ (convex combination):

    $$\min_w \sum_{j=1}^M \left( f_j^{(0)} - \sum_{k=1}^K f_j^{(k)} w_k \right)^2 ~,$$
    where $M$ is the number of variants, $K$ the number of populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the global population of interest.

---

## A set of global frequencies

<br>

- The 1000 Genomes data is too small, and missing some parts of the world

- HGDP, EGDP or SGDP are too small as well

--

<br>

- therefore I used the UK Biobank, which includes individuals from a bit of everywhere + secret recipe for ancestry estimation -> frequencies for 20 global populations

    <br>

    ```{r, out.width = "100%"}
    knitr::include_graphics("group-sizes.jpg")
    ```

---

## Verification

<br>

Computing $F_{ST}$ with closest population from the 1000G data:

<br>

```{r, out.width = "100%"}
knitr::include_graphics("group-fst.jpg")
```

---

## Controls (1/2)

<br>

- Biobank Japan:
    ```{r, out.width = "50%"}
    knitr::include_graphics("res-bbj.jpg")
    ```

<br>

- FinnGen:
    ```{r, out.width = "57%"}
    knitr::include_graphics("res-finngen.jpg")
    ```
---

## Controls (2/2)

<br>

- UK Biobank:
    ```{r, out.width = "60%"}
    knitr::include_graphics("res-ukbb.jpg")
    ```

---

## Other results (1/3)

- The COVID-19 Host Genetics Initiative ("European" subset):
    ```{r, out.width = "42%"}
    knitr::include_graphics("res-covid-eur.jpg")
    ```
- All (except 23andMe):
    ```{r, out.width = "48%"}
    knitr::include_graphics("res-covid-no23andme.jpg")
    ```

---

## Other results (2/3)

- Epilepsy (EUR + EAS + AFR):
    ```{r, out.width = "52%"}
    knitr::include_graphics("res-epilepsy.jpg")
    ```

<br>

> Furthermore, 531 cases of Asian descent, and 147 cases of African descent were included through a meta-analysis.

---

## Other results (3/3)

- PAGE consortium:
    ```{r, out.width = "45%"}
    knitr::include_graphics("res-page.jpg")
    ```

And the "truth" according to their paper:

```{r, out.width = "100%"}
knitr::include_graphics("page-truth.jpg")
```

---

## Conclusions

- Works well, but still not perfect (e.g. for single populations)

- Missing populations? (e.g. Native Americans, Native Hawaiian)

- Could add Ashkenazi Jewish group, but still a bit hesitant doing so

--

<br>

## Ideas for improvement

- Increase sample size of groups for more precise allele frequencies

- Use millions of variants (for now, only 500K genotyped variants of UKBB)

- Filter bad variants in the UKBB

- Give more weight to more ancestry-informative variants? to rarer variants?

### 2021-11-09 pres/group-meeting2.Rmd
save

---
title: "Group meeting #2"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 70)
knitr::opts_chunk$set(fig.align = 'center', dev = "svg", out.width = "70%",
                      echo = FALSE, comment = "", fig.width = 5, global.par = TRUE)
```

class: title-slide center middle inverse

<br>

# Using the UK Biobank as a global reference of worldwide populations:<br>application to measuring ancestry diversity from GWAS summary statistics

<br>

### Florian Privé

---

### Defining 26 diverse groups from the UKBB

```{r, out.width = "100%"}
knitr::include_graphics("admixture.jpg")
```

---

#### Minimum $F_{ST}$ between groups

```{r, out.width = "100%"}
knitr::include_graphics("min_fst.jpg")
```

---

### Inferring ancestry proportions from allele frequencies

I propose to find the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimize the following problem:
$$\sum_{j=1}^M w_j \left( f_j^{(0)} - \sum_{k=1}^K \alpha_k f_j^{(k)} \right)^2 ,$$
where $M=5.8 \cdot 10^6$ is the number of variants, $K=17$ the number of reference populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the cohort of interest.

--

***

This is similar to Summix (<small>Arriaga-MacKenzie, I. S., Matesi, G., Chen, S., Ronco, A., Marker, K. M., Hall, J. R., ... & Hendricks, A. E. (2021). Summix: A method for detecting and adjusting for population structure in genetic summary data. *The American Journal of Human Genetics*.</small>)

The difference: I introduce weights $w_j = 1 / \max\left(f_j^{(0)} (1 - f_j^{(0)}), ~0.01\right)$ to account for more variability in more common variants.

---

### Results

<br>

```{r, out.width = "100%"}
knitr::include_graphics("results-preprint.jpg")
```

---

### Alternative weights

<br>

- Accounting for LD? Using weights as inverse LD scores: $w_j = 1 / l_j$

--

- Accounting for more ancestry-informative variants?

    - getting truncated SVD $U \Delta V^T$ on $\frac{G_{i,j} - 2 f_j}{\sqrt{l_j}}$

    - using $w_j = \max_k{V_{j,k}}$ (also inversely correlated with $l_j$)

--

<br>

Based on the first tests I made, this basically made almost no difference to the results..

---

### Using penalization?

<br>

Solving instead (still $\alpha_k \ge 0$ and $\sum_k \alpha_k = 1$):

$$\sum_{j=1}^M w_j \left( f_j^{(0)} - \sum_{k=1}^K \alpha_k f_j^{(k)} \right)^2  - \lambda \sum_k \alpha_k^2 ~,$$

This tries to push the $\alpha_k$ to be as large as possible (but remember that they have to sum to one).

The maximum possible regularization is $\lambda_\text{max} = 0.99 \cdot \sigma_{17}$, where $\sigma_{17}$ is the minimum eigenvalue of $F$, where $F_{j,k} = \sqrt{w_j} f_j^{(k)}$.

[TODO: SHOW RESULTS]

---

### Alternative method: mixture of distributions

Consider the allele frequencies from a finite mixture distribution

$$p(x ~|~ \alpha) = \sum_{k=1}^K \alpha_k d_k(x) ~,$$
where $d_k(x) = dbeta\left(x,~ 1 + 2 \cdot N_k \cdot f^{(k)},~ 1 + 2 \cdot N_k \cdot (1 - f^{(k)}) \right)$ are the components densities (conjugate priors).

--

Then finding the maximum likelihood estimates of $\alpha$ can be solved my maximizing (still contrained):

$$\sum_{j=1}^M \log{\left( \sum_{k=1}^K \alpha_k L_j^{(k)} \right)},$$
where $L_j^{(k)} = d_k(f_j^{(0)})$.

This can be solved with R package mixsqp (<small>Kim, Y., Carbonetto, P., Stephens, M., & Anitescu, M. (2020). A fast algorithm for maximum likelihood estimation of mixture proportions using sequential quadratic programming. *Journal of Computational and Graphical Statistics*.</small>)

---

## Conclusions / Questions

<br>

- Was right to remove some groups? E.g. SE Europe. Could maybe merge South East and North East Europe

- Better Bayesian mixture to properly account for N? And allowing admixture

- Any other ideas to improve this? Either the ancestry groups, or the method to estimate ancestry proportions

### 2021-11-09 code/2-prepare-1000G-freq.R

@@ -1,74 +1,0 @@
-# https://www.cog-genomics.org/plink/2.0/resources#1kg_phase3
-pgen_zst <- runonce::download_file(
-  "https://www.dropbox.com/s/afvvf1e15gqzsqo/all_phase3.pgen.zst?dl=1",
-  dir = "tmp-data")
-pvar_zst <- runonce::download_file(
-  "https://www.dropbox.com/s/op9osq6luy3pjg8/all_phase3.pvar.zst?dl=1",
-  dir = "tmp-data")
-psam <- runonce::download_file(
-  "https://www.dropbox.com/s/yozrzsdrwqej63q/phase3_corrected.psam?dl=1",
-  dir = "tmp-data", fname = "all_phase3.psam")
-
-library(bigsnpr)
-plink2 <- download_plink2("tmp-data")
-
-sysglue <- function(...) system(glue::glue(...))
-pgen <- sub('\\.zst', '', pgen_zst)
-if (!file.exists(pgen)) sysglue("{plink2} --zst-decompress {pgen_zst} > {pgen}")
-pvar <- sub('\\.zst', '', pvar_zst)
-if (!file.exists(pvar)) sysglue("{plink2} --zst-decompress {pvar_zst} > {pvar}")
-
-# https://dx.doi.org/10.1016%2Fj.ajhg.2017.03.004
-outliers <- c("NA20314", "HG00731", "HG00732", "HG01880", "HG01882",
-              "HG01944", "HG02497", "NA20320", "NA20321")
-bigsnpr:::write.table2(cbind.data.frame(0, outliers), (tmp <- tempfile()))
-
-sysglue("{plink2} --pfile tmp-data/all_phase3",
-        " --remove {tmp} --keep-founders",
-        " --max-alleles 2 --mac 10 --autosome",
-        " --make-bed --out tmp-data/all_phase3",
-        " --threads {nb_cores()[[1]]} --memory 50e3")
-
-obj.bed <- bed("tmp-data/all_phase3.bed")
-
-library(doParallel)
-cl <- makeCluster(22)
-snp_info <- do.call("rbind", parLapply(cl, 1:22, function(chr) {
-  library(dplyr)
-  paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt") %>%
-    bigreadr::fread2(showProgress = FALSE) %>%
-    filter(V6 > 0.001, V8 > 0.3) %>%
[…]

### 2021-11-09 code/4-filter-merge-freq.R

@@ -65,7 +65,7 @@ print_fst <- function(pop, nb = 7) {
 }
 purrr::walk(names(pop_ukbb_size), ~ cat(print_fst(.), "\n"))
 # - [Japan] JPT: 0.00033 || CHB: 0.007 || CHS: 0.0089 || Asia (East): 0.01 || KHV: 0.014 || CDX: 0.017 || Philippines: 0.021
-# - [Asia (East)] CHS: 9e-04 || KHV: 0.0021 || CHB: 0.0029 || CDX: 0.0034 || Philippines: 0.01 || Japan: 0.01 || JPT: 0.011
+# - [Asia (East)] CHS: 0.0009 || KHV: 0.0021 || CHB: 0.0029 || CDX: 0.0034 || Philippines: 0.01 || Japan: 0.01 || JPT: 0.011
 # - [Philippines] Asia (East): 0.01 || KHV: 0.01 || CDX: 0.011 || CHS: 0.012 || CHB: 0.015 || Japan: 0.021 || JPT: 0.021
 # - [Africa (South)] Africa (Central): 0.0017 || LWK: 0.0035 || Caribbean: 0.0043 || Africa (West): 0.0044 || Africa (East 2): 0.005 || YRI: 0.005 || ESN: 0.0055
 # - [Africa (North)] PUR: 0.01 || Middle East: 0.01 || Italy: 0.01 || Europe (South West): 0.011 || TSI: 0.012 || IBS: 0.012 || Ashkenazi: 0.012

### 2021-11-09 pres/group-meeting2.Rmd
save

@@
 
-[TODO: SHOW RESULTS]
+---
+
+### Results of penalization (1/3)
+
+FinnGen sumstats:
+
+```{r, out.width = "75%"}
+knitr::include_graphics("regul-finngen.jpg")
+```
+
+---
+
+### Results of penalization (2/3)
+
+Epilepsy sumstats:
+
+```{r, out.width = "75%"}
+knitr::include_graphics("regul-epilepsy.jpg")
+```
+
+---
+
+### Results of penalization (3/3)
+
+PAGE sumstats:
+
+```{r, out.width = "75%"}
+knitr::include_graphics("regul-page.jpg")
+```
 
@@
 
+- Which alternative weights / method do you find the most interesting?
+
+- How to choose the best penalization? CV did not seem to work
+
 - Better Bayesian mixture to properly account for N? And allowing admixture

### 2021-11-30 code/4-filter-merge-freq.R
new code with proj

@@ -98,7 +98,34 @@ min_fst <- tibble::tibble(
   pop_ukbb = rownames(all_fst_1kg)[apply(all_fst_1kg, 2, which.min)],
   fst = apply(all_fst_1kg, 2, min)
 )
-print(arrange(min_fst, fst), n = Inf)
+print(dplyr::arrange(min_fst, fst), n = Inf)
+#    pop_1kg pop_ukbb                 fst
+#  1 JPT     Japan               0.000327
+#  2 STU     Sri Lanka           0.000340
+#  3 FIN     Finland             0.000411
+#  4 ACB     Caribbean           0.000483
+#  5 YRI     Africa (West)       0.000615
+#  6 BEB     Bangladesh          0.000675
+#  7 IBS     Europe (South West) 0.000794
+#  8 CHS     Asia (East)         0.000897
+#  9 CEU     United Kingdom      0.00101
+# 10 TSI     Italy               0.00108
+# 11 GBR     United Kingdom      0.00112
+# 12 ITU     Sri Lanka           0.00148
+# 13 ESN     Africa (West)       0.00149
+# 14 KHV     Asia (East)         0.00210
+# 15 MXL     South America       0.00235
+# 16 PJL     Pakistan            0.00260
+# 17 MSL     Africa (West)       0.00280
+# 18 CHB     Asia (East)         0.00288
+# 19 ASW     Caribbean           0.00295
+# 20 LWK     Africa (East 2)     0.00330
+# 21 CDX     Asia (East)         0.00337
+# 22 GIH     Sri Lanka           0.00426
+# 23 GWD     Africa (West)       0.00495
+# 24 CLM     South America       0.00527
+# 25 PUR     Europe (South West) 0.00941
+# 26 PEL     South America       0.0191
 
 all_scaled_fst <- purrr::pmap(min_fst, function(pop_1kg, pop_ukbb, fst) {
   var_fst <- snp_fst(list(
@@ -120,15 +147,16 @@ X[, "Finland"] <- (143 * X[, "Finland"] + 99 * X_1kg[ind, "FIN"]) / (143 + 99)
 X[, "Bangladesh"] <- (223 * X[, "Bangladesh"] + 86 * X_1kg[ind, "BEB"]) / (223 + 86)
 X[, "Japan"] <- (240 * X[, "Japan"] + 104 * X_1kg[ind, "JPT"]) / (240 + 104)
 X[, "South America"] <- (473 * X[, "South America"] + 84 * X_1kg[ind, "PEL"]) / (473 + 84)
[…]

### 2021-11-30 code/5-compute-PCA.R

@@ -1,0 +1,141 @@
+all_eid <- readRDS("tmp-data/info-UKBB.rds")$eid
+eid_sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+all_ind <- readRDS("data/list_ind_pop.rds")
+all_ind2 <- match(all_eid[unlist(all_ind)], eid_sample)
+
+all_freq <- readRDS("data/all_freq.rds")
+all_snp_id <- with(all_freq, split(paste(chr, pos, a0, a1, sep = "_"), as.factor(1:22)[chr]))
+
+library(bigsnpr)
+snp_readBGEN(
+  bgenfiles   = paste0("UKBB/bgen/ukb_imp_chr", 1:22, "_v3.bgen"),
+  list_snp_id = all_snp_id,
+  backingfile = "tmp-data/dosage_all_ind",
+  ind_row     = all_ind2,
+  ncores      = nb_cores()
+)
+
+ukbb <- snp_attach("tmp-data/dosage_all_ind.rds")
+G <- ukbb$genotypes
+CHR <- as.integer(ukbb$map$chromosome)
+POS <- ukbb$map$physical.pos
+ind_chr <- with(ukbb$map, split(seq_along(chromosome), chromosome))
+res_files <- paste0("tmp-data/ld_chr", 1:22, ".rds")
+
+if (all(file.exists(res_files))) {
+  future::plan("sequential")
+} else {
+  library(future.batchtools)
+  NCORES <- 15
+  plan(batchtools_slurm(workers = print(nrow(grid)), resources = list(
+    t = "12:00:00", c = NCORES, mem = "120g",
+    name = basename(rstudioapi::getSourceEditorContext()$path))))
+}
+
+ld <- unlist(furrr::future_map(1:22, function(chr) {
+  runonce::save_run({
+    ind <- ind_chr[[chr]]
+    bigsnpr::snp_ld_scores(G, ind.col = ind, size = 1000, infos.pos = POS[ind],
+                           ncores = NCORES)
+  }, file = res_files[chr])
+})) # 30 min for chr1 / 6 min for chr22
+
+
+## 1000G
+
+obj.bed <- bed("tmp-data/all_phase3.bed")
+fam <- dplyr::left_join(
+  obj.bed$fam,
+  bigreadr::fread2("https://figshare.com/ndownloader/files/31080292"))
[…]

### 2021-11-30 code/6-assess-ancestry-in-sumstats.R

@@ -4,7 +4,8 @@ library(dplyr)
 bigassertr::assert_dir("res_ancestry")
 
 all_freq <- readRDS("data/all_freq.rds")
-# bigreadr::fwrite2(mutate_at(all_freq, -(1:5), ~ round(., 5)), "ref_freqs.csv.gz")
+# bigreadr::fwrite2(all_freq, "ref_freqs.csv.gz")
+projection <- as.matrix(bigreadr::fread2("projection.csv.gz", select = paste0("PC", 1:16)))
 
 snp_match_ancestry <- function(sumstats, join_by_pos = TRUE) {
 
@@ -16,20 +17,13 @@ snp_match_ancestry <- function(sumstats, join_by_pos = TRUE) {
   ) %>%
     mutate(freq = ifelse(beta > 0, freq, 1 - freq))
 
-  X <- as.matrix(all_freq[matched$`_NUM_ID_`, -(1:5)])
-  y <- matched$freq
-  snp_ancestry_summary(y, X)
+  snp_ancestry_summary(
+    freq = matched$freq,
+    info_freq_ref = all_freq[matched$`_NUM_ID_`, -(1:5)],
+    projection = projection[matched$`_NUM_ID_`, ],
+    correction = c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099, 1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
+  )
 }
-
-
-## UKBB
-
-ukbb <- bed("../paper-ancestry-matching/data/ukbb.bed")
-af_UKBB <- bed_MAF(ukbb, ncores = nb_cores())$af
-sumstats <- transmute(ukbb$map, chr = chromosome, pos = physical.pos,
-                      a0 = allele2, a1 = allele1, freq = af_UKBB)
-str(res <- snp_match_ancestry(sumstats))
-saveRDS(res, "res_ancestry/UKBB.rds")
 
 
 ## T2D from Biobank Japan
@@ -46,7 +40,7 @@ sumstats <- fread2(
 ) %>%
   filter(chr %in% 1:22)
 
-str(res <- snp_match_ancestry(sumstats))
+(res <- snp_match_ancestry(sumstats))
 saveRDS(res, "res_ancestry/BBJ.rds")
 
 
@@ -59,7 +53,7 @@ sumstats <- bigreadr::fread2("../atw-pred/tmp-data/summary_stats_finngen_R5_T2D.
[…]

### 2021-11-30 tmp-code/test-regul-logLik.R

@@ -1,0 +1,168 @@
+coef <- 0.5
+p1 <- 0.2; N1 <- 200
+p2 <- 0.9; N2 <- 500
+hist(x1 <- rbeta(1e6, shape1.1 <- 2 * N1 * p1, shape2.1 <- 2 * N1 * (1 - p1)))
+hist(x2 <- rbeta(1e6, shape1.2 <- 2 * N2 * p2, shape2.2 <- 2 * N2 * (1 - p2)))
+hist(x <- coef * x1 + (1 - coef) * x2, "FD", freq = FALSE)
+(param <- EnvStats::ebeta(x)$param)
+curve(dbeta(x, param[1], param[2]), add = TRUE, col = "blue", lwd = 2)
+
+a <- coef * shape1.1 / (shape1.1 + shape2.1) + (1 - coef) * shape1.2 / (shape1.2 + shape2.2)
+c <- (1 - a) / a
+b <- coef^2 * (shape1.1 * shape2.1) / (shape1.1 + shape2.1)^2 / (shape1.1 + shape2.1 + 1) +
+  (1 - coef)^2 * (shape1.2 * shape2.2) / (shape1.2 + shape2.2)^2 / (shape1.2 + shape2.2 + 1)
+
+param1 <- (c / b / (1 + c)^2 - 1) / (1 + c)
+param2 <- c * param1
+curve(dbeta(x, param1, param2), add = TRUE, col = "green", lwd = 2)
+
+
+hist(rnorm(1e5))
+hist(0.5 * rnorm(1e5) + 0.5 * rnorm(1e5))
+
+###
+
+library(dplyr)
+
+all_freq <- bigreadr::fread2(
+  runonce::download_file("https://figshare.com/ndownloader/files/31218445",
+                         dir = "../bigsnpr/tmp-data", fname = "ref_freqs.csv.gz")
+)
+
+all_N <- bigreadr::fread2("all_prop.csv")$N
+all_N[c(8, 11, 12, 15)] <- all_N[c(8, 11, 12, 15)] + c(99, 84, 104, 86)
+
+
+# T2D from FinnGen
+(N <- 29193 + 182573)
+sumstats <- bigreadr::fread2("../freq-ancestry//tmp-data/summary_stats_finngen_R5_T2D.gz") %>%
+  transmute(rsid = rsids, chr = `#chrom`, pos, a0 = ref, a1 = alt,
+            freq = (2 * (n_hom_cases + n_hom_controls) +
[…]

### 2021-11-30 pres/group-meeting2.Rmd
typo

@@
 
-I propose to find the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimize the following problem:
+I propose to find the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimizes the following problem:
 $$\sum_{j=1}^M w_j \left( f_j^{(0)} - \sum_{k=1}^K \alpha_k f_j^{(k)} \right)^2 ,$$

### 2021-11-30 paper-ancestry-prop.tex
update text

--- 2021-10-27 paper-ancestry-prop.tex
@@
 	The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
-	Within this dataset, we carefully define 17 distinct ancestry groups from all four corners of the world. 
-	Using allele frequencies derived from these global reference groups, we are now able to effectively measure diversity from summary statistics of any genetic dataset.
+	Within this dataset, we carefully define 21 distinct ancestry groups from all four corners of the world. 
+	These ancestry groups can serve as a global reference of worldwide populations, with a handful of applications.
+	As an example application here, we use allele frequencies derived from these ancestry groups to effectively measure diversity from summary statistics of any genetic dataset.
 	Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
@@
 
-Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project, the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
-However, these datasets do not contain many individuals per population and therefore are not large enough to provide accurate allele frequency estimates for diverse worldwide populations.
-The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
+Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project (1KG), the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
+However, these datasets do not contain many individuals per population and therefore are not large enough for some purposes, such as accurately estimating allele frequency estimates for diverse worldwide populations.
+The UK Biobank (UKBB) project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
-Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 17 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-This includes five groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
-The detailed procedure used to define these groups is presented in the Supplementary Materials.
-We then provide reference allele frequencies for 5,816,590 genetic variants across these 17 diverse ancestry groups.
+Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 18 distinct ancestry groups from the UK Biobank to be used as global reference populations.
+This includes six groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
+The detailed procedure used to define these groups is presented in the first section of the Supplementary Materials.
 
-Then, as a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
-We propose to find the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimize the following problem:
-\(\sum_{j=1}^M \left(w_j  f_j^{(0)} - \sum_{k=1}^K \alpha_k w_j f_j^{(k)} \right)^2 ,\)
+Then we provide reference allele frequencies for 5,816,590 genetic variants across these 21 diverse ancestry groups (which are later merged in 18 groups in table \ref{tab:table1}).
+As a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
+\cite{arriaga2021summix} previously proposed method Summix, which finds the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimizes the following problem:
+\(\sum_{j=1}^M \left( f_j^{(0)} - \sum_{k=1}^K \alpha_k f_j^{(k)} \right)^2,\)
 where $M$ is the number of variants, $K$ the number of reference populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the cohort of interest.
-This is similar to Summix \cite[]{arriaga2021summix}, except for two differences.
-First, we introduce weights $w_j = 1 / \max\left(\sqrt{f_j^{(0)} (1 - f_j^{(0)})}, ~0.1\right)$ to account for more variability in more common variants.
-Second, and most importantly, we use the new reference set of allele frequencies we present here (instead of the five continental 1000 Genomes populations) for measuring genetic diversity with more precision.
-This optimization problem is a standard quadratic programming problem that can be solved using R package quadprog \cite[]{turlach2013quadprog}, and that we implement in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}.
+In \cite{arriaga2021summix}, they used the five continental 1KG populations as reference.
+Here, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals.
+This is described in more details in the second section of the Supplementary Materials.
+Finding the ancestry proportions in the PCA space (rather than using the allele frequencies directly) provides more power to distinguish between close populations. 
+This enable us to use more reference populations (18) in order to get a more fine-grained measure of genetic diversity.
+We implement this method in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}, relying on R package quadprog \cite[]{turlach2013quadprog} for solving this quadratic programming optimization problem.
 
-We download several GWAS summary statistics which reports allele frequencies and apply this method to both validate our new reference set we provide and show one of its potential use.
-We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirically validation.
-When applying this function (using the new reference set we provide) to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions estimates obtained match expectations (Table \ref{tab:table1}).
-When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 24.9\% South American, 21.9\% European, 35.4\% African, 2.0\% South Asian, 11.3\% East Asian, 3.9\% Filipino, and 0.5\% Middle East.
-GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented (Table \ref{tab:table1}).
+We download and apply this method to several GWAS summary statistics for which allele frequencies are reported.
+We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirically validation; when applying it to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions obtained match expectations (Table \ref{tab:table1}).
+When comparing our estimates with reported ancestries for more diverse cohorts, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 25.4\% South American, 22.6\% European (including 15.8\% from South-West Europe), 34.1\% African, 2.7\% South Asian, 10.6\% East Asian, and 4.6\% Filipino.
+GWAS summary statistics from either European ancestries or more diverse ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented (Table \ref{tab:table1}).
 
 % latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Wed Oct 27 13:40:20 2021
+% Tue Nov 30 16:24:09 2021
 \begin{table}[htb]
 	\centering
-	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c||c|c|c|c|c|c|c|c|}
+	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c|c|c|c|c|c|c|c|}
 		\hline
-		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
+		Population & N & BBJ & FinnGen & height-peru & arabic & t2d\_africa & GERA & PAGE & BrCa & PrCa & CAD & body-fat & covid & eczema & epilepsy & urate \\ 
 		\hline
-		Africa (West) & 735 &  &  &  &  & 36.7 & 1.4 & 1.9 & 26.4 & 0.2 & 0.2 & 2.4 & 1 & 4 & 0.3 & 0.6 & 3.6 \\ 
-		Africa (South) & 449 &  &  &  &  & 63.3 & 0.3 & 0.4 & 6.8 &  & 0.1 & 0.9 & 1.1 & 1.2 & 0.1 &  & 4 \\ 
-		Africa (East) & 276 &  &  &  & 11.4 &  &  &  & 0.2 &  &  &  & 0.2 & 0.1 &  &  & 1.4 \\ 
-		Africa (North) & 268 &  &  &  & 22.1 &  &  & 0.4 & 2 &  &  &  &  &  &  & 0.3 &  \\ 
+		Africa (West) & 735 &  &  &  &  & 30 & 1.9 & 27.7 & 0.3 & 0.3 & 2.2 & 0.7 & 4 & 0.2 & 0.7 & 2.2 \\ 
+		Africa (South) & 449 &  &  &  &  & 70 & 0.9 & 5.9 & 0.2 &  & 1.2 & 0.3 & 1.2 & 0.5 & 0.3 & 3.5 \\ 
+		Africa (East) & 276 &  &  &  & 13 &  &  &  &  &  & 0.1 &  & 0.3 &  &  & 1.9 \\ 
+		Africa (North) & 268 &  &  &  & 22 &  &  & 0.5 &  &  &  &  & 0.1 &  &  &  \\ 
 		\hline
-		Middle East & 523 &  &  &  & 63.5 &  & 1.3 & 3.4 & 0.5 & 2 & 2.2 & 5.2 & 2.9 & 1.5 & 1.6 & 3.2 & 2.8 \\ 
+		Middle East & 523 &  &  &  & 64.6 &  &  &  &  &  &  & 0.2 &  &  &  & 1.3 \\ 
 		\hline
-		Ashkenazi & 1975 &  &  &  & 1.9 &  & 1.4 & 6 & 0.9 & 1.3 & 3.3 & 2.2 & 2.3 & 1.3 & 2 & 3.6 & 1.5 \\ 
-		Europe (North East) & 320 &  & 8.9 &  &  &  & 3.8 & 7.7 & 0.7 & 9.6 & 10 & 10 & 11 & 11.7 & 10.1 & 8.9 & 7.5 \\ 
-		Finland & 143 (+ 99) &  & 85.2 &  &  &  & 2.7 & 3.5 & 0.7 & 11.9 & 14.8 & 8.2 & 10.5 & 15.4 & 13.1 & 7.1 & 4.3 \\ 
-		British-Irish Isles & 4000 &  & 5.1 &  &  &  & 79.1 & 53 & 8.9 & 59.8 & 56.1 & 46.6 & 49.2 & 54.8 & 61.7 & 62 & 36.6 \\ 
-		Europe (South West) & 603 &  &  &  &  &  & 7.7 & 12.2 & 10.7 & 11.2 & 11.2 & 7.5 & 9.5 & 5.3 & 10.1 & 11.2 & 7.3 \\ 
+		Ashkenazi & 1975 &  &  &  &  &  & 4.4 & 0.5 & 0.2 & 1.8 & 0.4 & 0.8 & 0.4 & 0.6 & 1.8 &  \\ 
+		Italy & 345 &  &  &  &  &  & 4.6 &  & 3.1 & 1.2 & 9.7 & 5.5 &  &  & 3.4 & 0.8 \\ 
+		Europe (East) & 667 &  &  &  &  &  & 10.5 &  & 6.9 & 11.3 & 10.5 & 11.4 & 13.2 & 11.7 & 13.9 & 10.8 \\ 
+		Finland & 143 (+ 99) &  & 100 &  &  &  & 2.4 & 0.7 & 9.7 & 13 & 5.9 & 8.8 & 14.8 & 12.8 & 6.5 & 2 \\ 
+		Europe (North West) & 4416 &  &  &  &  &  & 59.9 & 5.6 & 68.5 & 64.5 & 51.8 & 59.8 & 61.4 & 70.9 & 68 & 46 \\ 
+		Europe (South West) & 603 &  &  &  &  &  & 3.5 & 15.8 & 4.7 & 4.5 &  &  &  & 2.1 & 2.1 &  \\ 
 		\hline
-		South America & 473 (+ 84) &  & 0.4 & 99.1 &  &  & 0.1 & 4.5 & 24.9 & 0.9 & 0.6 & 1.4 & 0.7 & 1.8 & 0.4 & 0.1 & 0.2 \\ 
+		South America & 473 (+ 84) &  &  & 100 &  &  & 4.6 & 25.4 & 1.5 & 0.8 & 1.6 & 0.4 & 1.8 & 0.5 &  &  \\ 
 		\hline
-		Japan & 240 (+ 104) & 96.1 & 0.4 & 0.9 &  &  &  & 1.8 & 8.8 & 0.2 & 0.2 & 2.1 & 3 & 0.4 & 0.4 &  & 25.9 \\ 
-		Asia (East) & 961 & 3.9 &  &  &  &  & 0.2 & 3.7 & 2.5 & 0.6 & 0.3 & 2.5 & 1.8 & 0.2 &  & 3 & 2.3 \\ 
-		Philippines & 295 &  &  &  &  &  & 0.1 & 1.4 & 3.9 &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.5 \\ 
+		Sri Lanka & 372 &  &  &  & 0.4 &  & 0.4 &  & 3.4 & 1.7 & 4.7 & 4.2 & 1.8 &  &  & 1.9 \\ 
+		Pakistan & 400 &  &  &  &  &  &  & 1.1 &  &  & 7 & 4 &  &  &  &  \\ 
+		Bangladesh & 223 (+ 86) &  &  &  &  &  &  & 1.6 &  &  &  &  &  &  &  &  \\ 
 		\hline
-		Bangladesh & 223 (+ 86) &  &  &  &  &  & 0.6 &  & 1.2 & 1.1 & 0.8 & 2.8 & 2.8 & 0.9 &  &  & 1.4 \\ 
-		Sri Lanka & 372 &  &  &  & 1.1 &  & 0.4 &  & 0.8 & 0.8 & 0.2 & 2.2 & 1.6 & 0.4 &  &  & 0.7 \\ 
-		Pakistan & 400 &  &  &  &  &  & 0.8 &  &  & 0.4 &  & 6 & 2.5 & 0.7 &  &  &  \\ 
+		Asia (East) & 961 &  &  &  &  &  & 3.5 & 1.2 & 1.2 & 0.7 & 2.5 & 1.2 & 0.1 & 0.1 & 3.1 &  \\ 
+		Japan & 240 (+ 104) & 100 &  &  &  &  & 2.2 & 9.4 & 0.4 &  & 2.4 & 2.8 & 0.7 & 0.7 & 0.3 & 29.8 \\ 
+		Philippines & 295 &  &  &  &  &  & 1.5 & 4.6 &  &  &  &  & 0.2 &  &  &  \\ 
 		\hline
@@
 	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets.
-Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
+	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' were merged as a single ``Europe (North West)'' group.
+Similarly, ``Europe (North East)'' and ``Europe (South East)'' were merged as a single ``Europe (East)'' group.
+	Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
 Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
@@
 
-Here we provide an unprecedentedly large and diverse reference set of allele frequencies using the UK Biobank.
-We then show how to effectively measure diversity from GWAS summary statistics using this new set of reference allele frequencies.
+Here we identify an unprecedentedly large and diverse set of ancestry groups within a single cohort, the UK Biobank.
+Using allele frequencies derived from these ancestry groups (and the individual-level data for computing the PCs), we show how to effectively measure diversity from GWAS summary statistics reporting allele frequencies.
 Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
 Our work has limitations though. 
-First, although we managed to define ancestry groups with more than 240 individuals, it seems that these groups are still not large enough for the ancestry deconvolution method to recover 100\% of one ancestry (e.g.\ the case for the FinnGen and Biobank Japan cohorts which are likely very homogenous).
-In this case, we also capture a small proportion from nearby ancestry groups.
-Second, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
-Third, with the 17 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in both South Asia and South America.
-Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
-Anyway, despite these limitations, we envision that the new set of reference allele frequencies we provide will have many useful applications.
+First, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
+Second, with the 21 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in Europe, but we may still be missing more fine-grained diversity in other continents.
+Third, when we use the allele frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and, for meta-analyses of binary traits, whether they were computed as a weighted average of total or effective sample sizes.
+Despite these limitations, we envision that the ancestry groups we define here will have many useful applications.
+The presented method that uses these groups could e.g.\ be used to automatically report ancestry proportions in the GWAS Catalog \cite[]{macarthur2017new}.
+These ancestry groups could also be used for assigning ancestry in other cohorts using the PC projection from this study \cite[]{prive2021high}, assessing portability of polygenic scores \cite[]{prive2021high}, or deriving linkage disequilibrium (LD) references matching GWAS summary statistics from diverse ancestries.
 
@@
 The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-The set of reference allele frequencies for 5,816,590 genetic variants across 17 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31218445}.
+The set of reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31218445} [TODO: UPDATE].
 All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 

### 2021-12-01 paper-ancestry-prop-supp.tex
reexport code
save; almost finished

@@
 This enable us to identify variants with possible errors in the UK Biobank.
+We obtain a set of 5,816,590 variants left.
 
-\subsection{Designing the final set of reference populations}
+\subsection{Designing the final set of reference populations and preparing PCA projection}
 
@@
 We combine them with 2490 individuals from the 26 populations of the 1KG data, and use a pruned set of 252,893 high-quality genotyped variants to run an admixture analysis (K = 15) using function \texttt{snmf} from R package LEA \cite[]{frichot2014fast,gain2020lea}.
-Ancestry proportions for the 52 groups are represented in Figure \ref{fig:admixture}.
-We can see that we have defined very homogeneous groups; however based on these results and previous $F_{ST}$, we decide to slightly modify these groups.
-First, for the Finland, Bangladesh and Japan ancestry groups, we compute a weighted average of allele frequencies with the ones from the 1KG data (FIN, BEB and JPT) to increase the small sample size of these groups.
+Ancestry proportions for the 52 groups are represented in figure \ref{fig:admixture}.
+We can see that we have defined very homogeneous groups.
+Then, based on these results and previous $F_{ST}$, we merge four of the groups we defined with some 1KG populations.
+Indeed, for the ``Finland'', ``Bangladesh'' and ``Japan'' ancestry groups, we compute a weighted average of allele frequencies with the ones from the 1KG data (FIN, BEB and JPT) to increase the small sample size of these groups.
 We do the same for ``South America'' and ``PEL'' in order to capture slightly more of what is probably an Amerindian ancestral component (in blue).
-Second, we merge the ``Irish'' and ``United Kingdom'' groups as a new ``British-Irish Isles'' group.
-Third, we remove some groups that could be retrieved from combining two or three other groups: ``Africa (East 2)'', ``Caribbean'', ``Africa (Central)'', ``Scandinavia'', ``Italy'', ``Europe (South East)'', ``Europe (Central)'', and ``India''.
 
-\begin{figure}[hp]
-	\centerline{\includegraphics[width=0.9\textwidth]{admixture}}
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.92\textwidth]{admixture}}
 	\caption{Admixture coefficients (for K = 15 ancestral populations) for the individuals from the 26 ancestry groups we defined previously from the UK Biobank, as well as the 26 1KG populations.
@@
 
-This results in the set of 5,816,590 variants for 17 ancestry groups, which we provide for download and use to estimate ancestry proportions in GWAS summary statistics.
+We then compute the PCA from the 15,645 UKBB individuals and the 1KG individuals from the four added populations, across the 5,816,590 variants.
+We do not center the genotype matrix to capture one more component, and for an easier projection later.
+However, we do scale the genotypes by dividing them by the square root of LD scores to prevent from capturing LD structure \cite[]{zou2010quantification}.
+We verify that PC loadings are homogeneously distributed along the genome (no peak) in figure \ref{fig:loadings} and that PC scores are capturing population structure in  figure \ref{fig:scores} \cite[]{prive2020efficient}.
+Based on the PC scores and admixture results (Figures \ref{fig:admixture} and \ref{fig:scores}), we remove some groups that are admixed or too central, which therefore could be retrieved from combining two or three other groups: ``Africa (East 2)'', ``Caribbean'', ``Africa (Central)'', ``Europe (Central)'', and ``India''.
+This results in a set of 5,816,590 variants for 21 ancestry groups, which we provide for download and use to estimate ancestry proportions in GWAS summary statistics.
+We also provide PC loadings scaled so that they can be directly used with allele frequencies.
+Note that due to the shrinkage when projecting PCs, a correction needs to be applied when projecting new allele frequencies \cite[]{prive2020efficient}.
+We use six 1KG populations (YRI, IBS, STU, CHS, GBR, and TSI) and the corresponding UKBB ancestry groups we defined (``Africa (West)'', ``Europe (South West)'', ``Sri Lanka'', ``Asia (East)'', ``United Kingdom'', and ``Italy'') to estimate the correction coefficients.
+From PC1 to PC16, these are 1.000, 1.000, 1.000, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099, 1.123, 1.150, 1.195, 1.256, 1.321, 1.382, and 1.443.
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.92\textwidth]{all_loadings}}
+	\caption{PCA loadings over 5.8M variants for the PCA recomputed from the subset of UKBB and 1KG individuals used in the reference ancestry groups. \label{fig:loadings}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.85\textwidth]{all_scores}}
+	\caption{PCA scores for the PCA recomputed from the subset of UKBB and 1KG individuals used in the reference ancestry groups.
+Black points represent the projection of the allele frequencies from the 21 ancestry groups kept.
+\label{fig:scores}}
+\end{figure}
+
+Finally, the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) is estimated by minimizing the following problem:
+\(\sum_{l=1}^L \left( p_l^{(0)} - \sum_{k=1}^K \alpha_k p_l^{(k)} \right)^2,\)
+where $L$ is the number of PCs (16 here), $K$ the number of reference populations (21 here), $p_l^{(k)}$ is the projection of frequencies from population $k$ onto PC $l$, and $p_l^{(0)}$ is the (corrected) projection of frequencies from the cohort of interest onto PC $l$.
+Note that we replace the quadratic term from this problem by the nearest positive-definite matrix to make it solvable, because of the lower number of PCs compared to populations.
+Note that, because they are very close ancestry groups, we merge a posteriori the coefficients $\alpha_k$ from ``Ireland'', ``United Kingdom'' and ``Scandinavia'' into a single ``Europe (North West)'' group, and similarly for ``Europe (North East)'' and ``Europe (South East)'' into a single ``Europe (East)'' group.
 

### 2021-12-01 paper-ancestry-prop.tex
save; almost finished

@@
 This includes six groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
-The detailed procedure used to define these groups is presented in the first section of the Supplementary Materials.
+The detailed procedure used to define these groups is presented in the Supplementary Materials.
 
@@
 Here, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals.
-This is described in more details in the second section of the Supplementary Materials.
+This is described in more details in the Supplementary Materials.
 Finding the ancestry proportions in the PCA space (rather than using the allele frequencies directly) provides more power to distinguish between close populations. 
@@
 We download and apply this method to several GWAS summary statistics for which allele frequencies are reported.
-We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirically validation; when applying it to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions obtained match expectations (Table \ref{tab:table1}).
+We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirical validation; when applying it to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions obtained match expectations (Table \ref{tab:table1}).
 When comparing our estimates with reported ancestries for more diverse cohorts, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 25.4\% South American, 22.6\% European (including 15.8\% from South-West Europe), 34.1\% African, 2.7\% South Asian, 10.6\% East Asian, and 4.6\% Filipino.
@@
 		\hline
-		Population & N & BBJ & FinnGen & height-peru & arabic & t2d\_africa & GERA & PAGE & BrCa & PrCa & CAD & body-fat & covid & eczema & epilepsy & urate \\ 
+		Ancestry group & N & BBJ & FinnGen & Per\'u & Qatar & Africa & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
 		\hline
@@
 	\end{tabular}
-	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets.
+	\caption{Reference population with their size (N), and corresponding ancestry proportions (in \%) from several GWAS datasets.
 	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' were merged as a single ``Europe (North West)'' group.
@@
 The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-The set of reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31218445} [TODO: UPDATE].
+Refer to the help page of function \texttt{snp\_ancestry\_summary} for how to use it.
+The set of reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups defined here can be downloaded at \url{https://figshare.com/ndownloader/files/31620968} and PC loadings for all variants across 16 PCs at \url{https://figshare.com/ndownloader/files/31620953}.
 All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 

### 2021-12-01 code/6-assess-ancestry-in-sumstats.R

@@ -307,5 +307,5 @@ df <- bigreadr::fread2("all_prop.csv") %>%
 df %>%
   mutate_at(-(1:2), ~ ifelse(. == 0, "", as.character(.))) %>%
   xtable::xtable() %>%
-  print(caption.placement = "top", hline.after = c(-1, 0, 4, 5, 10, 11, 14, 17),
+  print(caption.placement = "top", hline.after = c(-1, 0, 4, 5, 11, 12, 15, 18),
         include.rownames = FALSE, file = "ancestry-proportions.tex")

### 2021-12-01 paper-ancestry-prop-supp.tex
finished

@@
 We do not center the genotype matrix to capture one more component, and for an easier projection later.
-However, we do scale the genotypes by dividing them by the square root of LD scores to prevent from capturing LD structure \cite[]{zou2010quantification}.
+However, we do scale the genotypes by dividing them by the square root of LD scores to prevent from capturing LD structure in the PCA \cite[]{zou2010quantification}.
 We verify that PC loadings are homogeneously distributed along the genome (no peak) in figure \ref{fig:loadings} and that PC scores are capturing population structure in  figure \ref{fig:scores} \cite[]{prive2020efficient}.

### 2021-12-01 paper-ancestry-prop.tex
finished

@@
 Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project (1KG), the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
-However, these datasets do not contain many individuals per population and therefore are not large enough for some purposes, such as accurately estimating allele frequency estimates for diverse worldwide populations.
+However, these datasets do not contain many individuals per population and therefore are not large enough for some purposes, such as accurately estimating allele frequencies for diverse worldwide populations.
 The UK Biobank (UKBB) project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
-Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 18 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-This includes six groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
+Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations.
+This includes nine groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
 The detailed procedure used to define these groups is presented in the Supplementary Materials.
 
-Then we provide reference allele frequencies for 5,816,590 genetic variants across these 21 diverse ancestry groups (which are later merged in 18 groups in table \ref{tab:table1}).
-As a use-case for the new reference set of allele frequencies we provide, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
+As an example application of these groups, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
 \cite{arriaga2021summix} previously proposed method Summix, which finds the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimizes the following problem:
@@
 In \cite{arriaga2021summix}, they used the five continental 1KG populations as reference.
-Here, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals.
+Here we provide reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups (which are later merged in 18 groups in table \ref{tab:table1}).
+Moreover, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals.
 This is described in more details in the Supplementary Materials.
 Finding the ancestry proportions in the PCA space (rather than using the allele frequencies directly) provides more power to distinguish between close populations. 
-This enable us to use more reference populations (18) in order to get a more fine-grained measure of genetic diversity.
+This enables us to use more reference populations (18) in order to get a more fine-grained measure of genetic diversity.
 We implement this method in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}, relying on R package quadprog \cite[]{turlach2013quadprog} for solving this quadratic programming optimization problem.
 
-We download and apply this method to several GWAS summary statistics for which allele frequencies are reported.
+We download several GWAS summary statistics for which allele frequencies are reported, and apply this method to them.
 We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirical validation; when applying it to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions obtained match expectations (Table \ref{tab:table1}).
@@
 	\end{tabular}
-	\caption{Reference population with their size (N), and corresponding ancestry proportions (in \%) from several GWAS datasets.
-	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' were merged as a single ``Europe (North West)'' group.
-Similarly, ``Europe (North East)'' and ``Europe (South East)'' were merged as a single ``Europe (East)'' group.
+	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) from several GWAS datasets.
+	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.
+Similarly, ``Europe (North East)'' and ``Europe (South East)'' are merged as a single ``Europe (East)'' group.
 	Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
-Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
+Several of these GWAS summary statistics have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}.
 \label{tab:table1}}
@@
 First, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
-Second, with the 21 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in Europe, but we may still be missing more fine-grained diversity in other continents.
-Third, when we use the allele frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and, for meta-analyses of binary traits, whether they were computed as a weighted average of total or effective sample sizes.
+Second, with the 21 ancestry groups we define, we think we are capturing a large proportion of the genetic diversity in Europe, but we may still be missing more fine-grained diversity in other continents.
+Third, when we use the allele frequencies reported in the GWAS summary statistics, it is not clear whether they were computed from all individuals (i.e.\ before performing any quality control and filtering), and, for meta-analyses of binary traits, whether they were computed as a weighted average of total or effective sample sizes.
 Despite these limitations, we envision that the ancestry groups we define here will have many useful applications.

### 2022-01-03 paper-ancestry-prop.tex
bioinfo

@@
 
-\noindent Contact:
-\begin{itemize}
-\item \url{florian.prive.21@gmail.com}
-\end{itemize}
-
 \vspace*{5em}
 
-\abstract{
-	The UK Biobank project is a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom.
-	Within this dataset, we carefully define 21 distinct ancestry groups from all four corners of the world. 
+\abstract{\noindent\textbf{Motivation:} Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.\\
+\textbf{Results:} Using the UK Biobank data, a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom, we carefully define 21 distinct ancestry groups from all four corners of the world.
 These ancestry groups can serve as a global reference of worldwide populations, with a handful of applications.
-	As an example application here, we use allele frequencies derived from these ancestry groups to effectively measure diversity from summary statistics of any genetic dataset.
-	Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
+Here we develop a method that use allele frequencies derived from these ancestry groups to effectively measure diversity from summary statistics of any genetic dataset.\\
+\textbf{Availability:} This method is implemented as function snp\_ancestry\_summary as part of R package bigsnpr.\\
+\textbf{Contact:} \href{florian.prive.21@gmail.com}{florian.prive.21@gmail.com}\\
+%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.
 }
@@
 
+\section*{Introduction}
+
 Several projects have focused on providing genetic data from diverse populations, such as the HapMap project, the 1000 genomes project (1KG), the Simons genome diversity project, and the human genome diversity project \cite[]{international2010integrating,10002015global,mallick2016simons,bergstrom2020insights}.
@@
 Despite being a cohort from the UK, this dataset is so large that it includes individuals that were born in all four corners of the world. 
+Therefore, the UK Biobank can serve as a global reference of worldwide populations when used in its entirety, i.e.\ without discarding valuable multi-ancestry genetic data.
+
+\section*{Implementation}
+
 Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-This includes nine groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
+These include nine groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
 The detailed procedure used to define these groups is presented in the Supplementary Materials.
-
-As an example application of these groups, we propose to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
+As a direct application of these groups, we propose a new method to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
 \cite{arriaga2021summix} previously proposed method Summix, which finds the convex combination of ancestry proportions $\alpha_k$ (positive and sum to 1) which minimizes the following problem:
@@
 
-We download several GWAS summary statistics for which allele frequencies are reported, and apply this method to them.
+\section*{Results}
+
+We download several GWAS summary statistics for which allele frequencies are reported, and apply this new method to them.
 We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirical validation; when applying it to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions obtained match expectations (Table \ref{tab:table1}).
@@
 
+\section*{Discussion}
+
 Here we identify an unprecedentedly large and diverse set of ancestry groups within a single cohort, the UK Biobank.
@@
 First, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
-Second, with the 21 ancestry groups we define, we think we are capturing a large proportion of the genetic diversity in Europe, but we may still be missing more fine-grained diversity in other continents.
-Third, when we use the allele frequencies reported in the GWAS summary statistics, it is not clear whether they were computed from all individuals (i.e.\ before performing any quality control and filtering), and, for meta-analyses of binary traits, whether they were computed as a weighted average of total or effective sample sizes.
+Second, with the 21 ancestry groups we define, we probably capture a large proportion of the genetic diversity in Europe, but more fine-grained diversity in other continents may still be lacking.
+Third, when using the allele frequencies reported in the GWAS summary statistics, it is not clear whether they were computed from all individuals (i.e.\ before performing any quality control and filtering), and, for meta-analyses of binary traits, whether they were computed as a weighted average of total or effective sample sizes.
 Despite these limitations, we envision that the ancestry groups we define here will have many useful applications.

### 2022-03-24 paper-ancestry-prop-supp.tex
add new results

@@
 \usepackage{makecell}
+\usepackage{placeins}
+\usepackage{attachfile}
 
@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 
+\FloatBarrier
+
+\subsection{Supplementary results}
+
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Thu Mar 24 14:00:51 2022
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c|c|c|c|c|c|c|c|}
+		\hline
+		Ancestry group & N & BBJ & FinnGen & Per\'u & Qatar & Africa & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
+		\hline
+		Africa (West) & 735 & 0.1 &  &  &  & 34.8 & 2.1 & 26.4 & 0.2 & 0.2 & 2.4 & 0.5 & 4 & 0.4 & 0.7 & 3.2 \\ 
+		Africa (South) & 449 &  &  &  &  & 65 & 0.5 & 6.9 & 0.2 & 0.1 & 1.1 & 0.5 & 1.3 & 0.1 & 0.1 & 3.4 \\ 
+		Africa (East) & 276 &  &  &  & 13 & 0.2 &  & 0.3 &  &  &  & 0.1 & 0.2 &  &  & 0.9 \\ 
+		Africa (North) & 268 &  &  &  & 22.5 &  & 0.1 & 1.8 &  &  &  &  &  &  &  &  \\ 
+		\hline
+		Middle East & 523 &  &  &  & 60.8 &  & 1 &  &  & 0.2 & 3 & 1.1 & 0.3 &  & 0.8 & 1 \\ 
+		\hline
+		Ashkenazi & 1975 &  &  &  & 1.5 &  & 5.1 & 0.4 & 0.5 & 2.5 & 1.3 & 1.6 & 0.6 & 0.9 & 2.4 & 0.8 \\ 
+		Italy & 345 &  &  &  &  &  & 5.1 & 2.4 & 2.7 & 3.3 & 3 & 3.8 & 0.8 & 3.4 & 5.6 & 2.5 \\ 
+		Europe (East) & 667 &  & 7.6 &  &  &  & 7.9 & 0.5 & 8.5 & 9.6 & 10.6 & 10.7 & 11.5 & 10.3 & 9.7 & 8.2 \\ 
+		Finland & 143 (+ 99) &  & 83.6 &  &  &  & 1.7 & 0.6 & 8.9 & 12.1 & 6.1 & 7.5 & 13 & 10 & 5.3 & 2.2 \\ 
+		Europe (North West) & 4416 &  & 7.4 &  &  &  & 57.8 & 9.5 & 67.7 & 63.7 & 51.7 & 57.2 & 60.8 & 69.4 & 66.4 & 43.8 \\ 
+		Europe (South West) & 603 &  &  &  &  &  & 6.8 & 9 & 6 & 5.8 & 3.3 & 4.3 & 2.3 & 4.3 & 5.2 & 3.6 \\ 
+		\hline
+		South America & 473 (+ 84) & 0.2 & 0.6 & 98.2 &  &  & 4.6 & 24.9 & 1.1 & 0.7 & 1.4 & 0.5 & 1.9 & 0.5 & 0.3 & 0.1 \\ 
+		\hline
+		Sri Lanka & 372 & 0.1 &  &  & 2.2 &  &  & 0.8 & 0.7 & 0.1 & 2.1 & 1.6 & 0.3 &  &  & 0.2 \\ 
+		Pakistan & 400 &  &  &  &  &  &  &  & 0.9 & 0.3 & 6.8 & 4.4 & 1.2 &  &  & 0.7 \\ 
+		Bangladesh & 223 (+ 86) & 0.2 &  &  &  &  & 0.3 & 1.3 & 1.4 & 0.9 & 2.9 & 2.4 & 0.9 &  &  & 0.7 \\ 
+		\hline
+		Asia (East) & 961 & 3.9 &  &  &  &  & 3.7 & 2.5 & 0.7 & 0.3 & 2.5 & 1.5 & 0.3 & 0.1 & 3.3 & 2.3 \\ 
+		Japan & 240 (+ 104) & 95.6 & 0.8 & 1.8 &  &  & 1.9 & 8.8 & 0.4 & 0.2 & 1.8 & 2.2 & 0.5 & 0.4 &  & 26.2 \\ 
+		Philippines & 295 &  &  &  &  &  & 1.4 & 4 &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.2 \\ 
+		\hline
+	\end{tabular}
+	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) inferred from the Summix method, for several GWAS summary statistics.
+		Note that an equivalent of the Summix method was used, relying on a different quadratic programming solver, since Summix did not converge properly when using 21 reference populations. 
+		Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.
+Similarly, ``Europe (North East)'' and ``Europe (South East)'' are merged as a single ``Europe (East)'' group.
+		Citations for the allele frequencies used are presented in Table 1 in the main text. \label{tab:summix}}
+\end{table}
+
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Thu Mar 24 14:18:58 2022
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c|c|c|c|c|c|c|c|}
+		\hline
+		Ancestry group & N & BBJ & FinnGen & Per\'u & Qatar & Africa & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
+		\hline
+		Africa (West) & 735 &  &  &  &  & 31.3 & 1.8 & 27.6 &  &  & 3.2 & 0.9 & 3.9 &  & 0.4 & 1.3 \\ 
+		Africa (South) & 449 &  &  &  &  & 68.7 & 1 & 5.9 & 0.4 & 0.4 &  & 0.1 & 1.3 & 0.7 &  & 4.4 \\ 
+		Africa (East) & 276 &  &  &  & 12.1 &  &  & 0.1 &  &  & 0.2 &  & 0.4 &  &  & 1.7 \\ 
+		Africa (North) & 268 &  &  &  & 23.8 &  &  &  &  &  &  &  & 0.3 &  & 2.1 &  \\ 
+		\hline
+		Middle East & 523 &  &  &  & 63.7 &  &  &  &  &  & 6.1 & 0.4 &  &  & 0.9 & 1.2 \\ 
+		\hline
+		Ashkenazi & 1975 &  &  &  &  &  & 4 & 0.1 &  & 2 & 0.6 & 0.7 & 0.7 & 0.5 & 2 &  \\ 
+		Italy & 345 &  &  &  &  &  & 3.8 &  & 6.5 & 0.7 &  & 8.5 &  &  &  & 3.6 \\ 
+		Europe (East) & 667 &  &  &  &  &  & 13.5 & 2.2 & 1.3 & 12.2 & 13.8 & 9 & 12.3 & 12.6 & 12.7 & 8.9 \\ 
+		Finland & 143 (+ 99) &  & 99.9 &  &  &  & 2.4 &  & 9.2 & 13.1 & 6.9 & 8.8 & 13.6 & 14.9 & 3 & 3.7 \\ 
+		Europe (North West) & 4416 &  &  &  &  &  & 57 & 4.1 & 67.5 & 63.9 & 54.5 & 60.3 & 62.2 & 67.9 & 65.5 & 44 \\ 
+		Europe (South West) & 603 &  &  &  &  &  & 4.4 & 16.8 & 8.5 & 4.5 &  &  &  & 2.5 & 9.9 &  \\ 
+		\hline
+		South America & 473 (+ 84) &  &  & 100 &  &  & 4.5 & 25.6 & 1.4 & 0.7 & 1.5 & 0.5 & 1.6 & 0.3 &  &  \\ 
+		\hline
+		Sri Lanka & 372 &  &  &  & 0.5 &  & 0.3 & 2.4 & 3.6 &  & 8.4 & 7 & 0.1 &  &  & 1.5 \\ 
+		Pakistan & 400 &  &  &  &  &  &  &  &  &  &  &  & 2.5 &  &  &  \\ 
+		Bangladesh & 223 (+ 86) &  &  &  &  &  &  &  &  & 1.9 &  &  &  &  &  &  \\ 
+		\hline
+		Asia (East) & 961 &  &  &  &  &  & 3.5 & 3.1 & 1.1 & 0.6 & 2.2 & 1.1 &  & 0.1 & 1.8 &  \\ 
+		Japan & 240 (+ 104) & 100 &  &  &  &  & 2.1 & 8.3 & 0.4 &  & 2.7 & 2.8 & 0.7 & 0.3 & 1.1 & 29.7 \\ 
+		Philippines & 295 &  & 0.1 &  &  &  & 1.6 & 3.7 &  &  &  &  & 0.5 & 0.3 & 0.5 &  \\ 
+		\hline
+	\end{tabular}
+	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) inferred from the proposed \texttt{snp\_ancestry\_summary} method while subsetting to 100,000 variants at random, for several GWAS summary statistics.
+	Note that an equivalent of the Summix method was used, relying on a different quadratic programming solver, since Summix did not converge properly when using 21 reference populations. 
+	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.
+Similarly, ``Europe (North East)'' and ``Europe (South East)'' are merged as a single ``Europe (East)'' group.
+	Citations for the allele frequencies used are presented in Table 1 in the main text. \label{tab:100K}}
+\end{table}
+
+\begin{table}[ht]
+	\centering
+	\attachfile[icon=Paperclip]{ancestry_sgdp.xlsx}{ double click on the icon to open the embedded Excel table}
+	\caption{Ancestry proportions (in \%) for all 345 individuals of the Simons genome diversity project \cite[]{mallick2016simons} inferred either from the proposed \texttt{snp\_ancestry\_summary} method (using the genotypes of an individual divided by 2 in place of allele frequencies) or using the projection analysis of ADMIXTURE (\texttt{-P}, \cite{shringarpure2016efficient}).
+	This table is also available in CSV format at \url{https://github.com/privefl/freq-ancestry/blob/main/ancestry_sgdp.csv}. \label{tab:sgdp}}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \clearpage

### 2022-03-24 paper-ancestry-prop.tex
add new results

@@
 	\end{tabular}
-	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) from several GWAS datasets.
+	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) inferred from the proposed \texttt{snp\_ancestry\_summary} method, for several GWAS summary statistics.
 	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.

### 2022-03-24 code/6-assess-ancestry-in-sumstats.R

@@ -4,8 +4,8 @@ library(dplyr)
 bigassertr::assert_dir("res_ancestry")
 
 all_freq <- readRDS("data/all_freq.rds")
-# bigreadr::fwrite2(all_freq, "ref_freqs.csv.gz")
-projection <- as.matrix(bigreadr::fread2("projection.csv.gz", select = paste0("PC", 1:16)))
+# fwrite2(all_freq, "ref_freqs.csv.gz")
+projection <- as.matrix(fread2("projection.csv.gz", select = paste0("PC", 1:16)))
 
 snp_match_ancestry <- function(sumstats, join_by_pos = TRUE) {
 
@@ -17,12 +17,56 @@ snp_match_ancestry <- function(sumstats, join_by_pos = TRUE) {
   ) %>%
     mutate(freq = ifelse(beta > 0, freq, 1 - freq))
 
-  snp_ancestry_summary(
+
+  freq_ref <- all_freq[matched$`_NUM_ID_`, -(1:5)]   #[c(1, 12, 15, 16, 19)]
+
+  res <- snp_ancestry_summary(
     freq = matched$freq,
-    info_freq_ref = all_freq[matched$`_NUM_ID_`, -(1:5)],
+    info_freq_ref = freq_ref,
     projection = projection[matched$`_NUM_ID_`, ],
-    correction = c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099, 1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
+    correction = c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099,
+                   1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
   )
+
+
+  # system.time(
+  #   res2 <- unlist(
+  #     Summix::summix(
+  #       data = cbind(freq = matched$freq, freq_ref),
+  #       reference = names(freq_ref),
+  #       observed = "freq"
+  #     )[names(freq_ref)]
+  #   )
+  # ) # > 3 min for 5 pops
+  # Summix v1.99.2 does not converge when using too many reference populations
+
+
+  # then this is equivalent to what Summix does (no PCA projection):
+  X <- as.matrix(freq_ref)
+  y <- matched$freq
+  # solve QP problem using https://stats.stackexchange.com/a/21566/135793
+  res3 <- quadprog::solve.QP(
[…]

### 2022-03-24 code/7-run-on-sgdp.R

@@ -1,0 +1,123 @@
+library(bigsnpr)
+library(dplyr)
+library(ggplot2)
+
+NCORES <- nb_cores()
+
+prefix <- "https://sharehost.hms.harvard.edu/genetics/reich_lab/sgdp/variant_set/cteam_extended.v4.maf0.1perc"
+bedfile <- runonce::download_file(paste0(prefix, ".bed"), dir = "tmp-data")
+runonce::download_file(paste0(prefix, ".fam"), dir = "tmp-data")
+bim_zip <- runonce::download_file(paste0(prefix, ".bim.zip"), dir = "tmp-data")
+unzip(bim_zip, exdir = "tmp-data", overwrite = FALSE)
+
+obj.bed <- bed(bedfile)
+dim(obj.bed)  # 345 x 34,418,131
+
+nbna_var <- bed_counts(obj.bed, ncores = NCORES)[4, ]
+mean(keep <- (nbna_var == 0))  # 88.6%
+
+all_freq <- bigreadr::fread2("ref_freqs.csv.gz")
+projection <- bigreadr::fread2("projection.csv.gz")
+correction <- c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099,
+                1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
+
+matched <- obj.bed$map %>%
+  transmute(chr = chromosome, pos = physical.pos, a0 = allele2, a1 = allele1,
+            beta = 1, keep) %>%
+  snp_match(all_freq[1:5]) %>%
+  filter(keep) %>%
+  select(-keep) %>%
+  print()
+# 3,545,628 variants left
+
+# project new individuals onto the PCA space
+all_proj <- sapply(1:16, function(k) {
+  bed_prodVec(
+    obj.bed,
+    projection[matched$`_NUM_ID_`, k + 5] / 2 * correction[k],
+    ind.col = matched$`_NUM_ID_.ss`,
+    # scaling to get G if beta = 1 and (2 - G) if beta = -1
+    center = 1 - matched$beta,
+    scale = matched$beta,
+    ncores = NCORES
+  )
+})
+
+# projected reference allele frequencies
+X <- crossprod(as.matrix(projection[matched$`_NUM_ID_`, -(1:5)]),
+               as.matrix(all_freq[matched$`_NUM_ID_`, -(1:5)]))
[…]

### 2022-03-25 paper-ancestry-prop-supp.tex
update text

@@
 
-In this section, we use the first 16 principal components (PCs, \cite{prive2020efficient}) reported by the UK Biobank (Data-Field 22009, \cite{bycroft2018uk}), the self-reported ancestry (Data-Field 21000), and the country of birth (Data-Field 20115).
+In this section, we use the principal components (PCs, Data-Field 22009), the self-reported ancestry (Data-Field 21000), and the country of birth (Data-Field 20115) from the UK Biobank \cite[]{bycroft2018uk} to first identify 26 homogeneous ancestry groups.
+We use the first 16 PCs only, which we have shown to be the PCs capturing population structure in the UK Biobank \cite[]{prive2020efficient}. 
+We compute Euclidean distances (L2-norms) from these PCs (i.e.\ in the PCA space), which we have shown to be an appropriate measure of the genetic distance between populations since these (squared) distances are largely proportional to the $F_{ST}$ between populations \cite[]{prive2021high}.
+
 For the individuals with a missing country of birth, we assign to them ``United Kingdom'' when they self-identify as ``British'' and ``Ireland'' when they self-identify as ``Irish''.
-Hereinafter, we use the Euclidean distance based on the first 16 PCs as we have shown to be an appropriate measure of the genetic distance between populations \cite[]{prive2021high}.
-The thresholds we use for this distance are based on carefully inspecting the self-reported ancestry and country of birth of individuals kept or excluded.
-
 Since the country of birth may poorly corresponds to the genetic ancestry that we are interested in, we perform some quality control on this information.
 Indeed, there are many individuals born in Africa with an Asian self-reported ancestry; we set their country of birth to missing. 
-We also set this information to missing for 1/ individuals born outside of Europe that are too close (log-distance lower than 4) to the UK; 2/ individuals born outside the UK within a log-distance of 2.5 to the UK; 3/ individuals born in Europe not within a log-distance of 5 to the UK.
-Finally, we keep only individuals born in South Africa with a log-distance to the UK larger than 6 (individuals of African ancestry). 
+We also set this information to missing for 1/ individuals born outside of Europe that are too close to the UK (log-distance to the UK center lower than 4); 2/ individuals born outside the UK within a log-distance of 2.5 to the UK; 3/ individuals born in Europe not within a log-distance of 5 to the UK.
+Finally, for individuals born in South Africa, we keep only the ones with a log-distance to the UK larger than 6 (individuals of African ancestry). 
+Note, for reference, that Ireland is at a log-distance of 3.3 from the UK, the Middle East at 4.7 and West Africa at 6. 
 After applying these filters, there are still many individuals born outside of the UK and Ireland, including e.g.\ 3352 individuals from India, 1100 from Germany, 1036 from Nigeria, and 76 countries with at least 50 individuals.
 
-Then, to define a homogeneous group to be used as a reference population, we choose a country, compute the robust PC center of individuals born in this country (using the geometric median, \cite{prive2021high}), and choose to include in this group all individuals within a chosen log-distance to the center.
-The distance threshold is chosen based on the countries of birth included.
+Then, to construct a homogeneous ancestry group to be used as a reference population, we choose a country, compute the robust PC center of individuals born in this country (using the geometric median, \cite{prive2021high}), and choose to include in this group all individuals within a chosen log-distance to the center.
+We investigate multiple choices for the countries and distance thresholds by carefully and manually inspecting the self-reported ancestry and country of birth of the individuals kept or excluded from these groups.
 For groups named after a country, we further restrict to individuals born in this country.
-We limit the ancestry groups to 2000 individuals in order not to have very uneven sample sizes; we pick at random among the United Kingdom and Ireland groups.
-We manually try different country centers and distance thresholds to cover as much of the worldwide populations as possible, without having overlapping groups.
+For the United Kingdom and Ireland groups, we pick 2000 individuals at random for each, in order not to have very uneven (larger) sample sizes compared to the other groups.
+We manually investigate different country centers and distance thresholds to cover as much of the worldwide populations as possible, without having overlapping groups.
 We also include two other groups: one ``Ashkenazi'' group based on a PC center we defined in \cite{prive2021high}, and a ``South America'' group based on individuals with $\text{PC6} > 50$ and $\text{PC8} < -30$ (clearly separated from the rest of individuals from the UK Biobank).
@@
 We also provide PC loadings scaled so that they can be directly used with allele frequencies.
-Note that due to the shrinkage when projecting PCs, a correction needs to be applied when projecting new allele frequencies \cite[]{prive2020efficient}.
+Note that, due to the shrinkage when projecting PCs, a correction needs to be applied when projecting new allele frequencies \cite[]{prive2020efficient}.
 We use six 1KG populations (YRI, IBS, STU, CHS, GBR, and TSI) and the corresponding UKBB ancestry groups we defined (``Africa (West)'', ``Europe (South West)'', ``Sri Lanka'', ``Asia (East)'', ``United Kingdom'', and ``Italy'') to estimate the correction coefficients.
@@
 Note that we replace the quadratic term from this problem by the nearest positive-definite matrix to make it solvable, because of the lower number of PCs compared to populations.
-Note that, because they are very close ancestry groups, we merge a posteriori the coefficients $\alpha_k$ from ``Ireland'', ``United Kingdom'' and ``Scandinavia'' into a single ``Europe (North West)'' group, and similarly for ``Europe (North East)'' and ``Europe (South East)'' into a single ``Europe (East)'' group.
+Note that, because they are very close ancestry groups, we merge a posteriori the ancestry coefficients $\alpha_k$ from ``Ireland'', ``United Kingdom'' and ``Scandinavia'' into a single ``Europe (North West)'' group, and similarly for ``Europe (North East)'' and ``Europe (South East)'' into a single ``Europe (East)'' group.
 
@@
 	Note that an equivalent of the Summix method was used, relying on a different quadratic programming solver, since Summix did not converge properly when using 21 reference populations. 
-		Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.
-Similarly, ``Europe (North East)'' and ``Europe (South East)'' are merged as a single ``Europe (East)'' group.
+	Note that, because they are very close ancestry groups, we merge a posteriori the ancestry coefficients $\alpha_k$ from ``Ireland'', ``United Kingdom'' and ``Scandinavia'' into a single ``Europe (North West)'' group, and similarly for ``Europe (North East)'' and ``Europe (South East)'' into a single ``Europe (East)'' group.
 	Citations for the allele frequencies used are presented in Table 1 in the main text. \label{tab:summix}}
@@
 	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) inferred from the proposed \texttt{snp\_ancestry\_summary} method while subsetting to 100,000 variants at random, for several GWAS summary statistics.
-	Note that an equivalent of the Summix method was used, relying on a different quadratic programming solver, since Summix did not converge properly when using 21 reference populations. 
-	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.
-Similarly, ``Europe (North East)'' and ``Europe (South East)'' are merged as a single ``Europe (East)'' group.
+	Note that, because they are very close ancestry groups, we merge a posteriori the ancestry coefficients $\alpha_k$ from ``Ireland'', ``United Kingdom'' and ``Scandinavia'' into a single ``Europe (North West)'' group, and similarly for ``Europe (North East)'' and ``Europe (South East)'' into a single ``Europe (East)'' group.
 	Citations for the allele frequencies used are presented in Table 1 in the main text. \label{tab:100K}}

### 2022-03-25 paper-ancestry-prop.tex
update text

@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=0.98\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 These ancestry groups can serve as a global reference of worldwide populations, with a handful of applications.
-Here we develop a method that use allele frequencies derived from these ancestry groups to effectively measure diversity from summary statistics of any genetic dataset.\\
+Here we develop a method that use allele frequencies and principal components derived from these ancestry groups to effectively measure ancestry proportions from allele frequencies of any genetic dataset.\\
 \textbf{Availability:} This method is implemented as function snp\_ancestry\_summary as part of R package bigsnpr.\\
@@
 
-Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations.
-These include nine groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (Table \ref{tab:table1}).
-The detailed procedure used to define these groups is presented in the Supplementary Materials.
+Here we carefully use information on self-reported ancestry, country of birth, and genetic similarity to define 21 distinct ancestry groups from the UK Biobank to be used as global reference populations, which is the first innovation of this paper.
+These include nine groups with genetic ancestries from Europe, four from Africa, three from South Asia, three from East Asia, one from the Middle East, and one from South America (which are later merged into 18 groups in table \ref{tab:table1}).
+The detailed procedure used to construct these reference ancestry groups is presented in the Supplementary Materials.
 As a direct application of these groups, we propose a new method to estimate global ancestry proportions from a cohort based on its allele frequencies only (i.e.\ summary statistics).
@@
 where $M$ is the number of variants, $K$ the number of reference populations, $f_j^{(k)}$ is the frequency of variant $j$ in population $k$, and $f_j^{(0)}$ is the frequency of variant $j$ in the cohort of interest.
-In \cite{arriaga2021summix}, they used the five continental 1KG populations as reference.
-Here we provide reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups (which are later merged in 18 groups in table \ref{tab:table1}).
+\cite{arriaga2021summix} used the five continental 1KG populations as reference.
+Here we provide reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups (which are later merged into 18 groups in table \ref{tab:table1}).
 Moreover, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals.
-This is described in more details in the Supplementary Materials.
-Finding the ancestry proportions in the PCA space (rather than using the allele frequencies directly) provides more power to distinguish between close populations. 
-This enables us to use more reference populations (18) in order to get a more fine-grained measure of genetic diversity.
-We implement this method in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}, relying on R package quadprog \cite[]{turlach2013quadprog} for solving this quadratic programming optimization problem.
+Instead, we then minimize 
+\(\sum_{l=1}^L \left( p_l^{(0)} - \sum_{k=1}^K \alpha_k p_l^{(k)} \right)^2,\)
+with the same convex constraints and where $L$ is the number of PCs, $p_l^{(k)}$ is the projection of allele frequencies from population $k$ onto PC $l$, and $p_l^{(0)}$ is the (corrected) projection of allele frequencies from the cohort of interest onto PC $l$.
+Finding the ancestry proportions in the PCA space (rather than using the allele frequencies directly) provides more power to distinguish between close populations, which is the second innovation of this paper.
+This enables us to use more reference populations in order to get a more fine-grained measure of genetic diversity.
+We implement this method in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}.
+The steps required by the proposed method are then 1/ read all summary statistics datasets into R; 2/ match variants and alleles between summary statistics and the reference allele frequencies we provide; 3/ project allele frequencies onto the PCA space (matrix multiplication); 4/ solve the final (small) quadratic programming problem, by relying on R package quadprog \cite[]{turlach2013quadprog}.
+All these steps are very fast and overall require a few minutes only for GWAS summary statistics with millions of variants.
 
@@
 	\caption{Reference populations with their size (N), and corresponding ancestry proportions (in \%) inferred from the proposed \texttt{snp\_ancestry\_summary} method, for several GWAS summary statistics.
-	Note that ``Ireland'', ``United Kingdom'' and ``Scandinavia'' are merged as a single ``Europe (North West)'' group.
-Similarly, ``Europe (North East)'' and ``Europe (South East)'' are merged as a single ``Europe (East)'' group.
+	Note that, because they are very close ancestry groups, we merge a posteriori the ancestry coefficients $\alpha_k$ from ``Ireland'', ``United Kingdom'' and ``Scandinavia'' into a single ``Europe (North West)'' group, and similarly for ``Europe (North East)'' and ``Europe (South East)'' into a single ``Europe (East)'' group.
 	Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}.
@@
 
+We then perform three secondary analyses.
+First, we compare the results obtained previously in Table \ref{tab:table1} with the results we would get without using the PCA projection of allele frequencies (i.e.\ equivalent to the Summix method).
+The resulting ancestry proportions are presented in Table S1 and are clearly less precise for Biobank Japan (BBJ) and FinnGen.
+Second, we compare previous results with the ones obtained using a smaller number of variants, by randomly sampling 100,000 variants to run the proposed method.
+The resulting ancestry proportions are presented in Table S2 and are highly consistent with the ones from Table \ref{tab:table1}, showing that 100,000 overlapping variants are enough to run the proposed method.
+Third, we also infer ancestry proportions for all 345 individuals of the Simons genome diversity project \cite[]{mallick2016simons} using the reference allele frequencies we provide and two methods.
+We use either our proposed method with the genotypes of an individual divided by 2 in place of allele frequencies, or by using the projection analysis of ADMIXTURE (\texttt{-P}, \cite{shringarpure2016efficient}).
+Results are very consistent between the two methods, and are overall as expected, further validating the proposed ancestry groups and the proposed method to infer ancestry proportions, which seems very precise even at the individual level.
+
 \section*{Discussion}
 
-Here we identify an unprecedentedly large and diverse set of ancestry groups within a single cohort, the UK Biobank.
-Using allele frequencies derived from these ancestry groups (and the individual-level data for computing the PCs), we show how to effectively measure diversity from GWAS summary statistics reporting allele frequencies.
+Here we have identified an unprecedentedly large and diverse set of ancestry groups within a single cohort, the UK Biobank.
+Using allele frequencies and principal components (PCs) derived from these ancestry groups, we show how to effectively measure diversity from GWAS summary statistics reporting allele frequencies.
 Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
 Our work has limitations though. 
-First, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
+First, it is unknown whether we can effectively capture any existing ancestry as a combination of the 21 reference populations we defined.
+For example, it seems that Native Hawaiians in the PAGE study are partly captured by the ``Philippines'' ancestry group we define.
 Second, with the 21 ancestry groups we define, we probably capture a large proportion of the genetic diversity in Europe, but more fine-grained diversity in other continents may still be lacking.
@@
 
-The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
+The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}) and a recent enough version can be installed from CRAN.
 Refer to the help page of function \texttt{snp\_ancestry\_summary} for how to use it.
@@
 F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath) and by a Lundbeck Foundation Fellowship (R335-2019-2339 to Bjarni J. Vilhj\'almsson).
+The author thanks Shai Carmi for helpful discussions.
 The author thanks GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
@@
 
-The author declares no competing interests.
+The author does not have any competing interest to declare.
 

### 2022-03-28 paper-ancestry-prop.tex
typo

@@
 These ancestry groups can serve as a global reference of worldwide populations, with a handful of applications.
-Here we develop a method that use allele frequencies and principal components derived from these ancestry groups to effectively measure ancestry proportions from allele frequencies of any genetic dataset.\\
+Here we develop a method that uses allele frequencies and principal components derived from these ancestry groups to effectively measure ancestry proportions from allele frequencies of any genetic dataset.\\
 \textbf{Availability:} This method is implemented as function snp\_ancestry\_summary as part of R package bigsnpr.\\

### 2022-05-11 paper-ancestry-prop.tex
minor text changes

@@
 \usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
+\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
 \usepackage{rotating}
@@
 
-\usepackage{xr}
-\externaldocument{paper-ancestry-prop-supp}
+%\usepackage{xr}
+%\externaldocument{paper-ancestry-prop-supp}
 
@@
 
-\abstract{\noindent\textbf{Motivation:} Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.\\
+\begin{abstract}
+\noindent\textbf{Motivation:} Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.\\
 \textbf{Results:} Using the UK Biobank data, a prospective cohort study with deep genetic and phenotypic data collected on almost 500,000 individuals from across the United Kingdom, we carefully define 21 distinct ancestry groups from all four corners of the world.
@@
 %\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.
-}
+\end{abstract}
 
@@
 \cite{arriaga2021summix} used the five continental 1KG populations as reference.
+
 Here we provide reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups (which are later merged into 18 groups in table \ref{tab:table1}).
-Moreover, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals.
+Moreover, we rely on the projection of our reference allele frequencies onto the PCA (principal component analysis) space computed from the corresponding UKBB (and 1KG) individuals, and also make these PC loadings available for download.
 Instead, we then minimize 
 \(\sum_{l=1}^L \left( p_l^{(0)} - \sum_{k=1}^K \alpha_k p_l^{(k)} \right)^2,\)
-with the same convex constraints and where $L$ is the number of PCs, $p_l^{(k)}$ is the projection of allele frequencies from population $k$ onto PC $l$, and $p_l^{(0)}$ is the (corrected) projection of allele frequencies from the cohort of interest onto PC $l$.
+with the same convex constraints on ancestry proportions $\alpha_k$, and where $L$ is the number of PCs, $p_l^{(k)}$ is the projection of allele frequencies from population $k$ onto PC $l$, and $p_l^{(0)}$ is the (corrected) projection of allele frequencies from the cohort of interest onto PC $l$.
+Note that we need to correct for the shrinkage when projecting a new dataset (here the allele frequencies from the GWAS summary statistics) onto the PC space \cite[]{prive2020efficient}.
 Finding the ancestry proportions in the PCA space (rather than using the allele frequencies directly) provides more power to distinguish between close populations, which is the second innovation of this paper.
 This enables us to use more reference populations in order to get a more fine-grained measure of genetic diversity.
-We implement this method in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}.
-The steps required by the proposed method are then 1/ read all summary statistics datasets into R; 2/ match variants and alleles between summary statistics and the reference allele frequencies we provide; 3/ project allele frequencies onto the PCA space (matrix multiplication); 4/ solve the final (small) quadratic programming problem, by relying on R package quadprog \cite[]{turlach2013quadprog}.
+
+The steps required by the proposed method are then 1/ read all summary statistics datasets into R, i.e.\ the reference allele frequencies and corresponding PC loadings we provide for download as well as the GWAS summary statistics containing the allele frequencies of interest; 2/ match variants and alleles between summary statistics and the reference allele frequencies we provide; 3/ project allele frequencies onto the PCA space (matrix multiplication); 4/ solve the final (small) quadratic programming problem, by relying on R package quadprog \cite[]{turlach2013quadprog}.
+Steps 3 and 4 are now implemented in function \texttt{snp\_ancestry\_summary} in our R package bigsnpr \cite[]{prive2017efficient}.
+Step 2 can be performed using existing function \texttt{snp\_match}.
+A tutorial is provided at \url{https://privefl.github.io/bigsnpr/articles/ancestry.html}.
 All these steps are very fast and overall require a few minutes only for GWAS summary statistics with millions of variants.
@@
 The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}) and a recent enough version can be installed from CRAN.
-Refer to the help page of function \texttt{snp\_ancestry\_summary} for how to use it.
+A tutorial on ancestry proportions and ancestry grouping is available at \url{https://privefl.github.io/bigsnpr/articles/ancestry.html}.
 The set of reference allele frequencies for 5,816,590 genetic variants across 21 diverse ancestry groups defined here can be downloaded at \url{https://figshare.com/ndownloader/files/31620968} and PC loadings for all variants across 16 PCs at \url{https://figshare.com/ndownloader/files/31620953}.
